---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sharkABC3D

<!-- badges: start -->
<!-- badges: end -->

The goal of sharkABC3D is to facilitate a range of analyses for querying abiotic factors by shark species. 

## Functionality
- Querying of World Ocean Atlas 2023 for marine species ranges, in terms of area and depth ranges.

## Installation

You can install the development version of extractclimate3d from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Marine-Biodiversity-Conservation-Lab/extractclimate3d")
```

Make sure to add your IUCN API key for accessing Red List data. Sign up and acquire your API key here: https://api.iucnredlist.org/ 

```{r}
usethis::edit_r_environ()
# Add the following line to the .Renviron file
IUCN_API_KEY="your_iucn_api_key_here"
```