---
title: "Visualizing 2.5D Species-Fishery Overlap in a Raster Cell"
output: html_document
---

<!--
This is NOT a package vignette. It is the source that builds the conceptual
schematic for the Bangladesh analysis (figures/fig-overlap-schematic.png). It
lives in figures/ and is excluded from the built package via .Rbuildignore.
Render it to regenerate the figure (ggsave writes into figures/ whether run from
the project root or here); it depends on tidyterra, patchwork, and sf, plus the
GEBCO and World EEZ files referenced in the Bangladesh vignette.

Supersedes an earlier hexagonal-grid version of this schematic. The analysis it
illustrates is vignettes/bangladesh-fisheries-3d-overlap.Rmd.
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Visualizing 2.5D Species-Fishery Overlap in a Raster Cell

The Bangladesh analysis
(`vignettes/bangladesh-fisheries-3d-overlap.Rmd`) measures the overlap between a
species and a fishery as a **volume**, one grid cell at a time. This document
builds the two-panel schematic that explains how, with a synthetic species and
fishery so the geometry is easy to read.

The representation is **2.5D**, not a stack of cubes: the study area is
discretized *horizontally* into square cells, but each cell stores a single
**continuous** depth interval per feature. Volumes are exact interval
arithmetic, not a count of occupied depth bins.

- **Panel A** — the study area. The Bangladesh EEZ on a Lambert Azimuthal Equal
  Area grid, each cell carrying its seafloor depth from GEBCO. An inset zooms in
  far enough to resolve individual square cells and highlights the one Panel B
  dissects.
- **Panel B** — that single cell's water column, drawn in isometric projection:
  the species depth interval, the fishery depth interval, and the intersection
  that `calc_volume_overlap()` returns.

The three quantities Panel B shows are exactly the ones the analysis computes:

| Panel B element | Analysis quantity |
|---|---|
| Purple slab | `depth_min` / `depth_max` from `vect_to_envelope()` for the species |
| Yellow slab | Same, for the fishery (after the 3 m tidal correction) |
| Teal slab | `depth_min_overlap` / `depth_max_overlap` from `calc_volume_overlap()` |
| Slab footprint | Cell area from `terra::cellSize()` |
| Blue floor | Seafloor depth, which truncates every slab from below |

## Setup

```{r setup}
# devtools::load_all()
library(sharkabc3d)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(patchwork)

sf::sf_use_s2(FALSE)
```

## Panel A: the study area grid

We rebuild the study grid the way the Bangladesh vignette does, so the schematic
describes the real analysis rather than an idealisation of it: the same GEBCO
release, the same EEZ polygon, and the same equal-area projection. The one
deliberate departure is the cell size — the analysis runs at 1000 m, but the
figure uses a coarser grid so that individual cells are visible in the inset.
That is a legibility choice, not a change to the method.

```{r study-grid}
bathy <- load_gebco_bathymetry(
  "~/Programming_Projects/Big_Data/gebco_2025_sub_ice_topo/GEBCO_2025_sub_ice.nc"
)

bangladesh_eez <- sf::st_read(
  "~/Programming_Projects/Big_Data/World_EEZ_v12_20231025/eez_v12.shp",
  query = "SELECT * FROM eez_v12 WHERE TERRITORY1 = 'Bangladesh'",
  quiet = TRUE
) %>% rename(geometry = "_ogr_geometry_")

# Lambert Azimuthal Equal Area centred on Bangladesh waters: this is what makes
# every cell the same size on the ground, and so what licenses treating
# the cell area as a constant when converting depth intervals to volumes.
bangladesh_crs <- "+proj=laea +lat_0=22 +lon_0=90 +datum=WGS84 +units=m"

# Grid resolution in metres. The Bangladesh analysis runs at 1000 m; this figure
# is free to use a coarser grid, because at 1 km the cells are far too small to
# resolve in the inset. Everything downstream — the highlighted cell footprint,
# the inset window, the cell area used for the volume arithmetic, and the panel
# labels — derives from this one number, so changing it here keeps the whole
# figure self-consistent.
grid_res <- 10000

study_grid <- create_study_raster(
  layers = list(bangladesh_eez),
  res = grid_res,
  crs = bangladesh_crs
)

# Crop before projecting so we never reproject the full global GEBCO grid.
# terra::project() resamples bilinearly by default, so each cell takes an
# interpolated seafloor depth rather than the mean of the ~450 m GEBCO cells it
# contains. clamp(lower = 0) puts land at zero depth.
bathy_crop <- terra::crop(bathy, terra::vect(bangladesh_eez))
seafloor <- terra::project(bathy_crop, study_grid)
seafloor <- terra::clamp(-seafloor, lower = 0)

# Mask to the EEZ: the analysis extent is the EEZ polygon, not the grid's
# bounding box, and the figure should show the same footprint.
eez_proj <- sf::st_transform(bangladesh_eez, bangladesh_crs)
seafloor_eez <- terra::mask(seafloor, terra::vect(eez_proj))
names(seafloor_eez) <- "depth"

# Read the realised cell size back off the raster rather than trusting
# `grid_res`: create_study_raster() fits whole cells to the layer extent, so the
# result can differ slightly from the requested resolution. These drive the
# highlighted cell, the inset window, and every "N km" label in the figure.
cell_dx <- terra::res(seafloor_eez)[1]
cell_dy <- terra::res(seafloor_eez)[2]

# Label for the cell size. Written as km when the cells are at least 1 km, and
# dropped to a whole number when it is one (so "10 km", not "10.0 km").
cell_label <- paste0(round(cell_dx / 1000, digits = 0), " km")
cell_area_km2 <- (cell_dx / 1000) * (cell_dy / 1000)

c(cell_dx = cell_dx, cell_dy = cell_dy, cell_area_km2 = cell_area_km2)
```

### Choosing the cell to dissect

Panel B needs one cell, and the two panels should agree: the seafloor depth
drawn in Panel B is read from the very cell the inset highlights. We pick a
mid-shelf cell — deep enough that a fishery does not reach the seabed, shallow
enough that the species' nominal lower limit gets truncated — because that is
the case worth explaining.

```{r focus-cell}
# A point on the mid shelf, in lon/lat, projected onto the study grid.
focus_lonlat <- sf::st_sfc(sf::st_point(c(90.5, 20.1)), crs = 4326)
focus_xy <- sf::st_coordinates(sf::st_transform(focus_lonlat, bangladesh_crs))

focus_cell <- terra::cellFromXY(seafloor_eez, focus_xy)
focus_centre <- terra::xyFromCell(seafloor_eez, focus_cell)
focus_depth <- terra::extract(seafloor_eez, focus_cell)[["depth"]]

# Round for the schematic: this is a conceptual figure, not a data readout.
focus_depth <- round(focus_depth / 10) * 10
focus_depth
#> [1] 450        # at grid_res = 10000; ~400 m at the analysis's 1000 m

# Footprint of the highlighted cell, as a polygon for both the inset and the
# leader line from Panel A to Panel B. Built from half a cell either side of the
# cell centre, so it tracks the grid resolution instead of assuming one.
focus_poly <- terra::as.polygons(
  terra::ext(
    focus_centre[1] - cell_dx / 2, focus_centre[1] + cell_dx / 2,
    focus_centre[2] - cell_dy / 2, focus_centre[2] + cell_dy / 2
  )
) |> sf::st_as_sf()
sf::st_crs(focus_poly) <- bangladesh_crs
```

### The full-extent map

At the full extent of the EEZ the cells are at or below pixel size, so we draw
the seafloor raster directly rather than thousands of cell polygons.
`coord_sf(datum = ...)` keeps the plot in the equal-area projection while
drawing a familiar lon/lat graticule.

```{r map-full}
# Box marking the region the inset magnifies. Sized in CELLS, not metres, so the
# inset always shows a legible number of them however the grid is resolved.
inset_cells <- 12
inset_ext <- terra::ext(
  focus_centre[1] - inset_cells / 2 * cell_dx,
  focus_centre[1] + inset_cells / 2 * cell_dx,
  focus_centre[2] - inset_cells / 2 * cell_dy,
  focus_centre[2] + inset_cells / 2 * cell_dy
)
inset_box <- sf::st_as_sf(terra::as.polygons(inset_ext))
sf::st_crs(inset_box) <- bangladesh_crs

depth_fill <- scale_fill_gradientn(
  name = "Seafloor\ndepth (m)",
  colours = c("#bcdcf5", "#4191d2", "#2166ac", "#0a2850"),
  na.value = "transparent",
  trans = "reverse"          # shallow pale at the top of the legend
)

p_map <- ggplot() +
  geom_sf(data = eez_proj, fill = "grey97", colour = "grey60",
          linewidth = 0.3, linetype = "dashed") +
  geom_spatraster(data = seafloor_eez) +
  depth_fill +
  geom_sf(data = inset_box, fill = NA, colour = "grey15", linewidth = 0.5) +
  geom_sf(data = focus_poly, fill = NA, colour = "#21918c", linewidth = 0.6) +
  coord_sf(datum = sf::st_crs(4326), expand = FALSE) +
  labs(title = paste0("A. Study area: ", cell_label, " equal-area grid")) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    axis.title = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "inside",
    legend.position.inside = c(0.82, 0.18),
    legend.background = element_rect(fill = "white", colour = "grey70"),
    legend.key.height = unit(0.9, "cm")
  )
p_map
```

### The inset: individual cells

Zoomed to a window a dozen cells across, the cells resolve. Here we *do* draw them as
polygons, with their boundaries visible, because the whole point of the inset is
that the grid is square and discrete — the single change from the earlier
hexagonal version of this analysis.

```{r map-inset}
inset_rast <- terra::crop(seafloor_eez, inset_ext)

# One polygon per cell, boundaries drawn. aggregate = FALSE keeps cells separate
# instead of dissolving equal-valued neighbours into one shape.
inset_cells <- terra::as.polygons(inset_rast, aggregate = FALSE, na.rm = TRUE) |>
  sf::st_as_sf()

p_inset <- ggplot() +
  geom_sf(data = inset_cells, aes(fill = depth),
          colour = "grey20", linewidth = 0.08) +
  depth_fill +
  geom_sf(data = focus_poly, fill = NA, colour = "#3ce8dc", linewidth = 1.1) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "none",
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8)
  )
p_inset
```

## Panel B: the water column of one cell

Panel B is an isometric drawing, built from ggplot polygons rather than a
rendered 3D scene. Two reasons: the slabs are translucent and deliberately
interpenetrating, which path tracers render unpredictably; and a polygon plot
exports as vector art, so the callout labels stay editable when the figure is
finished in Inkscape.

### Isometric projection helpers

An isometric view maps a 3D point to the page by shearing the two horizontal
axes apart and letting depth run straight up and down. Increasing `x` moves
right-and-back, increasing `y` moves left-and-back, and `z` (negative below the
sea surface) moves straight down. The viewer therefore sits over the
low-`x`, low-`y` corner and sees each box's top face plus its two near sides.

```{r iso-helpers}
# Camera. The azimuth stays at 45 degrees, so the two horizontal axes are
# treated symmetrically and the cell reads as a square seen corner-on. The
# elevation is how far the eye sits ABOVE the sea-surface plane: 90 degrees
# would look straight down (the plane fills the page, no side faces visible),
# and 0 would put the eye exactly in the plane (the plane collapses to a line
# and only the side faces show). Lowering it flattens the water surface and
# swings the view toward side-on, which is what makes the depth intervals — the
# thing this panel is about — read more clearly than a true isometric does.
VIEW_AZIMUTH_DEG <- 45
VIEW_ELEV_DEG    <- 30

az   <- VIEW_AZIMUTH_DEG * pi / 180
elev <- VIEW_ELEV_DEG * pi / 180

ISO_C <- cos(az)              # horizontal spread of the x/y axes
ISO_S <- sin(az) * sin(elev)  # how much horizontal depth projects up the page

# Page units per metre of depth. The cell footprint is 1 unit wide while the
# column is hundreds of metres deep, so depth has to be compressed hard or the
# drawing becomes a needle.
#
# Both panels share ONE scale, keyed to the deeper of the two cells, so the
# columns are directly comparable: Panel C's shallower cell must actually look
# shorter than Panel B's. Scaling each panel to its own depth would draw them
# the same height and quietly hide the difference the pair exists to show.
# Tall on purpose. The panels are laid out in a row, so horizontal space is the
# scarce dimension and vertical space is nearly free — exaggerating depth uses
# that space, separates the slabs, and gives the callouts a long span to spread
# down. This is a drawing scale only: the depth NUMBERS in the labels are the
# real ones, and no volume is computed from the page geometry.
COLUMN_HEIGHT <- 6.5          # page height of the DEEPEST column drawn

# Panel C uses a hardcoded shallow bed (see that section): declared here because
# the shared vertical scale has to know about it.
shallow_seafloor_m <- 200

VSCALE <- COLUMN_HEIGHT / max(focus_depth, shallow_seafloor_m)

iso <- function(x, y, z) {
  data.frame(
    px = (x - y) * ISO_C,
    py = (x + y) * ISO_S + z * VSCALE
  )
}

# Darken a colour toward black; used to shade the three visible faces of a box
# so it reads as a solid rather than a flat silhouette.
darken <- function(col, f) {
  grDevices::rgb(t(grDevices::col2rgb(col) * f), maxColorValue = 255)
}

# The three visible faces of an axis-aligned box, as a polygon data frame.
# z_top and z_bot are depths in metres below the surface (positive down); the
# box spans from z_top down to z_bot.
iso_box <- function(x0, x1, y0, y1, z_top, z_bot, colour) {
  zt <- -z_top
  zb <- -z_bot

  face <- function(pts, shade, id) {
    d <- iso(pts[, 1], pts[, 2], pts[, 3])
    d$grp <- id
    d$fill <- darken(colour, shade)
    d
  }

  rbind(
    # Top face: brightest.
    face(cbind(c(x0, x1, x1, x0), c(y0, y0, y1, y1), rep(zt, 4)), 1.00, "top"),
    # y = y0 face, falling away to the right of the viewer.
    face(cbind(c(x0, x1, x1, x0), rep(y0, 4), c(zb, zb, zt, zt)), 0.82, "right"),
    # x = x0 face, falling away to the left of the viewer.
    face(cbind(rep(x0, 4), c(y0, y1, y1, y0), c(zb, zb, zt, zt)), 0.64, "left")
  )
}

# Draw one box as three shaded polygons at a fixed transparency.
geom_iso_box <- function(box, alpha = 0.55) {
  geom_polygon(
    data = box,
    aes(x = px, y = py, group = grp, fill = fill),
    alpha = alpha, colour = "grey25", linewidth = 0.2
  )
}
```

### The synthetic depth ranges

The depths below are invented for legibility; only the seafloor depth is real,
read from the highlighted cell in Panel A. They are chosen to show the two
behaviours that matter in the analysis:

1. The species' nominal lower limit is compared against the seafloor, which
   `vect_to_envelope()` takes as one more `depth_max` constraint, so the bed
   wins wherever it is shallower than the stated limit. Whether that
   bites depends on the cell: this one is ~450 m, so a 250 m limit is **not**
   clamped. Raise `species_bot_nominal` past the seafloor depth to show the
   truncation, which is the behaviour the panel is meant to illustrate.
2. The fishery works the full water column down to the bed, so it encloses the
   species' whole depth range. The intersection is therefore the entire species
   interval: this cell offers **no depth refuge**, and the species is fully
   exposed to the gear where the two footprints meet.

```{r depth-ranges}
seafloor_m <- focus_depth        # from Panel A's highlighted cell (450 m here)

species_top <- 80
species_bot_nominal <- 250                            # IUCN / study depth limit
species_bot <- min(species_bot_nominal, seafloor_m)   # clamped by vect_to_envelope()

fishery_top <- 0
fishery_bot <- 120

# The intersection calc_volume_overlap() computes: interval arithmetic on the
# two already-clamped intervals.
overlap_top <- max(species_top, fishery_top)
overlap_bot <- min(species_bot, fishery_bot)
overlap_thickness <- max(0, overlap_bot - overlap_top)

c(species = species_bot - species_top,
  fishery = fishery_bot - fishery_top,
  overlap = overlap_thickness)
#>  species  fishery  overlap
#>      170      120       40

# Volumes, in the km^3 the analysis reports: the cell's footprint area times the
# interval thickness expressed in km. cell_area_km2 comes from the study grid
# above, so these track the chosen resolution.
c(species_km3 = cell_area_km2 * (species_bot - species_top) / 1000,
  overlap_km3 = cell_area_km2 * overlap_thickness / 1000)
```

### Assembling the scene

The three slabs share the cell's footprint but are offset horizontally so
all three stay visible — an exploded view, not a claim that they occupy
different ground. The overlap slab is drawn last and least transparent, since it
is the quantity the figure exists to explain.

```{r iso-scene}
col_species   <- "#7b52ab"   # purple, matching the vignette's overlap maps
col_fishery   <- "#e8d54a"   # yellow
col_intersect <- "#21918c"   # teal
col_water     <- "#7fb2e5"
col_bed       <- "#3b5aa0"

# One transparency for every depth band. Applied once per band (see the banding
# in column_panel), so the whole column reads at a consistent weight.
SLAB_ALPHA <- 0.65

# How far the species and fishery slabs are pulled apart. The offsets must be
# OPPOSED in x and y: this projection has px = (x - y) * ISO_C, so shifting both
# axes by the same amount moves a box straight up the page and separates nothing.
# Shifting -x/+y sends the species left, +x/-y sends the fishery right, and both
# stay at the same height because py depends on (x + y), which is unchanged.
#
# At off = 0 all three slabs share one footprint exactly, which is geometrically
# truthful but hides the intersection inside the other two. Raise this to pull
# the species left and the fishery right and expose the overlap slab between.
off <- 0

# How far the overlap slab is inset from the cell footprint on every side.
# At 0 it is drawn flush, filling the cell exactly as the species and fishery do
# — which is the truthful footprint, since the intersection spans the whole cell
# horizontally and differs only in depth. The cost is that its near faces land on
# top of the fishery's in the overlap band, so the fishery reads as stopping at
# the top of the overlap rather than continuing to its own lower limit. Raise
# this to nest the overlap inside the others and keep their outlines continuous.
overlap_inset <- 0

# The cell spans 0..1 in both horizontal axes, so its middle is at 0.5. The sea
# surface is centred there rather than on the origin, so the plane sits squarely
# around the column instead of hanging off one corner of it.
cell_mid <- 0.5

# Half-width of the sea-surface plane, in cell widths either side of the middle.
surf <- 1.5

# Shared drawing limits for every column panel. coord_equal() fits a panel's own
# data range into the space it is given, so a panel whose range is half as tall
# gets scaled up twice as much — which is why Panel C's grid cell rendered LARGER
# than Panel B's despite both being a 1 x 1 footprint. Pinning both panels to one
# set of limits, sized to the deepest column drawn, makes the cell identical in
# every panel and lets the shorter column simply leave space below it.
PANEL_DEPTH_MAX <- max(focus_depth, shallow_seafloor_m)

panel_px <- 2 * surf * ISO_C                       # half-width of the water plane
panel_py_top <- 2 * (cell_mid + surf) * ISO_S      # far corner of the water plane
panel_py_bot <- -PANEL_DEPTH_MAX * VSCALE - 0.03   # underside of the deepest bed

PANEL_XLIM <- c(-panel_px - 0.15, panel_px + 0.15)
PANEL_YLIM <- c(panel_py_bot - 0.35, panel_py_top + 0.35)

# All callouts stack in ONE column to the right of the drawing. With the panels
# in a row there is ample vertical space but little horizontal, so a single
# right-hand stack keeps every label inside its own panel instead of overrunning
# into the neighbouring one.
label_x <- 2.4

# Spread the labels evenly down that column, ordered by the height of the
# feature each one points at, so leader lines do not cross. The span adapts to
# the panel's own geometry, which matters because Panel C's column is shorter
# than Panel B's under the shared vertical scale.
spread_labels <- function(ty, top, bot) {
  out <- numeric(length(ty))
  out[order(-ty)] <- seq(top, bot, length.out = length(ty))
  out
}

# Build one complete column panel. Every depth is in metres below the surface.
# `species_bot_nominal` is the species' stated lower limit BEFORE clamping, so
# the function can report truthfully whether vect_to_envelope() actually
# truncated it at this cell's bed.
column_panel <- function(seafloor_m, species_top, species_bot_nominal,
                         fishery_top, fishery_bot, title, cell_label) {

  # Clamp exactly as vect_to_envelope() does, then intersect the two clamped
  # intervals exactly as calc_volume_overlap() does.
  species_bot <- min(species_bot_nominal, seafloor_m)
  fishery_bot <- min(fishery_bot, seafloor_m)
  was_clamped <- species_bot_nominal > seafloor_m

  overlap_top <- max(species_top, fishery_top)
  overlap_bot <- min(species_bot, fishery_bot)
  has_overlap <- overlap_bot > overlap_top

  # Partition the column into DISJOINT depth bands and draw each exactly once.
  # The three slabs share one footprint, so drawing them as three overlapping
  # boxes STACKS their alpha: the band covered by species + fishery + overlap
  # composites three translucent layers and renders near-opaque, while a band
  # covered by only one stays light. That is why the transparency looked uneven
  # between the top, middle, and bottom of the column. Since the slabs are flush,
  # only the outermost surface is ever visible anyway — so banding is visually
  # equivalent and gives every band the same, uniform alpha.
  edges <- sort(unique(c(species_top, species_bot, fishery_top, fishery_bot)))
  bands <- data.frame(top = head(edges, -1), bot = tail(edges, -1))
  bands$mid <- (bands$top + bands$bot) / 2

  # Which features contain each band. Teal where both do (that IS the overlap),
  # otherwise the single feature's colour; bands inside neither are dropped.
  in_sp <- bands$mid > species_top & bands$mid < species_bot
  in_fi <- bands$mid > fishery_top & bands$mid < fishery_bot
  bands$fill <- ifelse(in_sp & in_fi, col_intersect,
                ifelse(in_sp, col_species,
                ifelse(in_fi, col_fishery, NA)))
  bands <- bands[!is.na(bands$fill) & bands$bot > bands$top, , drop = FALSE]

  # Drawn shallowest first, so each band's top face paints over the one above it
  # and reads as the boundary between them.
  band_layers <- lapply(seq_len(nrow(bands)), function(i) {
    geom_iso_box(iso_box(0, 1, 0, 1, bands$top[i], bands$bot[i], bands$fill[i]),
                 alpha = SLAB_ALPHA)
  })

  # Sea surface: a translucent plane at z = 0, centred on the cell's middle.
  water_plane <- iso(cell_mid + c(-surf, surf, surf, -surf),
                     cell_mid + c(-surf, -surf, surf, surf), rep(0, 4))

  # The cell's seafloor: a thin plate at the bed, capping the column from below.
  # Its thickness is set in page units and converted back to metres, so it stays
  # the same visual thickness at any depth rather than vanishing on deep cells.
  bed_plate <- iso_box(0, 1, 0, 1,
                       seafloor_m, seafloor_m + 0.03 / VSCALE, col_bed)

  # Guide lines dropping from the cell corners to the bed, so the column the
  # slabs belong to is unambiguous.
  corners <- expand.grid(x = c(0, 1), y = c(0, 1))
  guides <- do.call(rbind, lapply(seq_len(nrow(corners)), function(i) {
    d <- iso(rep(corners$x[i], 2), rep(corners$y[i], 2), c(0, -seafloor_m))
    d$grp <- i
    d
  }))

  # Leader targets, computed through iso() so they stay pinned to the geometry.
  at <- function(x, y, depth) iso(x, y, -depth)
  sp_left_x <- -off;   sp_mid_y <- off + 0.5
  fi_right_y <- -off;  fi_mid_x <- off + 0.5

  targets <- rbind(
    surface = at(cell_mid - surf * 0.8, cell_mid + surf * 0.8, 0),
    sp_top  = at(sp_left_x, sp_mid_y, species_top),
    sp_bot  = at(sp_left_x, sp_mid_y, species_bot),
    fi_top  = at(fi_mid_x, fi_right_y, fishery_top),
    fi_bot  = at(fi_mid_x, fi_right_y, fishery_bot),
    bed     = at(0, 0, seafloor_m),
    overlap = at(0, 0, if (has_overlap) (overlap_top + overlap_bot) / 2 else NA)
  )

  labels <- c(
    surface = "Sea surface (0 m)",
    sp_top  = paste0("Upper depth of species (", species_top, " m)"),
    sp_bot  = if (was_clamped) {
      paste0("Lower depth of species\nclamped to seafloor (", species_bot, " m)")
    } else {
      paste0("Lower depth of species (", species_bot, " m)")
    },
    fi_top  = paste0("Upper depth of fishery (", fishery_top, " m)"),
    fi_bot  = paste0("Lower depth of fishery (", fishery_bot, " m)"),
    bed     = paste0("Seafloor depth of grid cell (", seafloor_m, " m)"),
    overlap = if (has_overlap) {
      paste0("Overlap volume (", overlap_top, "-", overlap_bot, " m)")
    } else NA_character_
  )

  callouts <- data.frame(
    key = rownames(targets),
    label = labels[rownames(targets)],
    tx = targets$px,
    ty = targets$py,
    stringsAsFactors = FALSE
  )
  callouts <- callouts[!is.na(callouts$label), ]

  # Vertical span the labels are spread across: the geometry's own extent, with
  # a little headroom so the topmost label clears the water plane.
  py_all <- c(water_plane$py, guides$py, bed_plate$py)
  top <- max(py_all) + 0.35
  bot <- min(py_all) - 0.35

  callouts$lx <- label_x
  callouts$hjust <- 0
  callouts$ly <- spread_labels(callouts$ty, top, bot)

  ggplot() +
    # Water surface, drawn first so everything else sits over it.
    geom_polygon(data = water_plane, aes(px, py),
                 fill = col_water, alpha = 0.28, colour = NA) +
    # Column guide lines from the cell corners down to the bed.
    geom_path(data = guides, aes(px, py, group = grp),
              colour = "grey45", linewidth = 0.25, linetype = "22") +
    geom_iso_box(bed_plate, alpha = 0.95) +
    # The species / overlap / fishery bands, every one at the same alpha.
    band_layers +
    scale_fill_identity() +
    # Callout leader lines, then the labels over them. geom_label (rather than
    # annotate) so label.size = NA is honoured and the boxes stay borderless.
    geom_segment(data = callouts,
                 aes(x = lx, xend = tx, y = ly, yend = ty),
                 colour = "grey30", linewidth = 0.3) +
    geom_label(data = callouts,
               aes(x = lx, y = ly, label = label, hjust = hjust),
               size = 4.1, label.size = NA, fill = "white", alpha = 0.85,
               lineheight = 0.95) +
    coord_equal(xlim = PANEL_XLIM, ylim = PANEL_YLIM, clip = "off") +
    labs(title = title) +
    theme_void(base_size = 13) +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = 15),
      # Nearly all the slack goes to the right, where the label stack lives.
      plot.margin = margin(10, 150, 10, 15)
    )
}
```

### Panel B: the deep cell

The cell Panel A highlights sits in deep water, so the species' 250 m lower
limit is shallower than the bed and `vect_to_envelope()` leaves it untouched. The
fishery stops well above the seafloor, leaving a thick depth refuge below it.

```{r panel-b}
p_column <- column_panel(
  seafloor_m          = seafloor_m,
  species_top         = species_top,
  species_bot_nominal = species_bot_nominal,
  fishery_top         = fishery_top,
  fishery_bot         = fishery_bot,
  title               = "B. Deep cell: species sits clear of the seafloor",
  cell_label          = cell_label
)
p_column
```

### Panel C: a shallow cell, where clamping bites

Panel B alone does not show the behaviour that most shapes the Bangladesh
results. On the shallow continental shelf the seafloor is often shallower than a
species' stated lower depth limit, and `vect_to_envelope()` truncates the range at
the bed — which is exactly why this package's volumes come out below the
original hexagonal-grid analysis (see the vignette's Step 4).

Panel C makes that case explicit with a **hardcoded 200 m bed**. It is not read
from the bathymetry: the point is to hold the species and fishery identical to
Panel B and vary only the seafloor, so the difference between the panels is
attributable to the clamp and nothing else. Against a 200 m bed the same 250 m
species limit *is* truncated, and the fishery reaches the seabed.

```{r panel-c}
p_column_shallow <- column_panel(
  seafloor_m          = shallow_seafloor_m,
  species_top         = species_top,
  species_bot_nominal = species_bot_nominal,
  fishery_top         = fishery_top,
  fishery_bot         = fishery_bot,
  title               = "C. Shallow cell: species range clamped to the seafloor",
  cell_label          = cell_label
)
p_column_shallow
```

## Composing the figure

The map carries the inset in its upper-right corner via
`patchwork::inset_element()`, then the three panels run left to right in a
single row. B and C share one vertical scale, so the shallower cell in C reads
as genuinely shorter than B rather than merely relabelled — which is easier to
judge side by side than stacked.

```{r compose-figure, fig.width = 20, fig.height = 8}
p_map_inset <- p_map +
  inset_element(p_inset, left = 0.52, bottom = 0.62, right = 0.99, top = 0.99)

fig_overlap_schematic <- (p_map_inset | p_column | p_column_shallow) +
  plot_layout(widths = c(1.1, 1, 1))
fig_overlap_schematic

# Always write into the figures/ directory, whether render runs from the project
# root or from figures/ itself.
fig_dir <- if (basename(getwd()) == "figures") "." else "figures"
ggsave(
  filename = file.path(fig_dir, "fig-overlap-schematic_raw.pdf"),
  plot = fig_overlap_schematic,
  width = 20, height = 8, dpi = 300, bg = "white"
)
ggsave(
  filename = file.path(fig_dir, "fig-overlap-schematic_raw.png"),
  plot = fig_overlap_schematic,
  width = 20, height = 8, dpi = 300, bg = "white"
)
```

`fig-overlap-schematic_raw.pdf` is the raw export; hand-edit it in Inkscape to
produce the final `fig-overlap-schematic.pdf` / `.png` used in the manuscript.
Callout labels are spread automatically down each side of a panel by
`spread_labels()`, so they rarely collide. The knobs most likely to want nudging
in Inkscape are `label_x` (how far the stack sits from the drawing),
`COLUMN_HEIGHT` (vertical exaggeration), and the view angle.

## Recap

- The grid is **square** and equal-area — the change from the earlier
  hexagonal version of this analysis.
- Each cell holds one **continuous** depth interval per feature — a
  `SpatEnvelope`. Nothing is snapped to standard depth levels; the vertical
  dimension is analytical, and only the horizontal is discretized. That is what
  makes the model 2.5D rather than a true voxel stack.
- `vect_to_envelope()` takes the seafloor as one more `depth_max` constraint, so
  every interval is truncated at the cell's bed before any overlap is computed.
  Whether that truncation is visible in Panel B depends on the focus cell: it
  bites only when the nominal limit is deeper than that cell's bed.
- `calc_volume_overlap()` intersects the two clamped intervals and multiplies
  the resulting thickness by the cell area. Summed over cells, that is the
  overlap volume the analysis reports.
- The gap between the teal slab and the seafloor is the **depth refuge**: water
  the species occupies that the fishery does not reach.
