---
title: "Visualizing 2.5D Species-Fishery Overlap in a Raster Cell"
output: html_document
---

<!--
This is NOT a package vignette. It is the source that builds the conceptual
schematic for the Bangladesh analysis (figures/fig-overlap-schematic.png). It
lives in figures/ and is excluded from the built package via .Rbuildignore.
Render it to regenerate the figure (ggsave writes into figures/ whether run from
the project root or here); it depends on tidyterra, patchwork, and sf, plus the
GEBCO and World EEZ files referenced in the Bangladesh vignette.

Supersedes an earlier hexagonal-grid version of this schematic. The analysis it
illustrates is vignettes/bangladesh-fisheries-3d-overlap.Rmd.
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Visualizing 2.5D Species-Fishery Overlap in a Raster Cell

The Bangladesh analysis
(`vignettes/bangladesh-fisheries-3d-overlap.Rmd`) measures the overlap between a
species and a fishery as a **volume**, one grid cell at a time. This document
builds the two-panel schematic that explains how, with a synthetic species and
fishery so the geometry is easy to read.

The representation is **2.5D**, not a stack of cubes: the study area is
discretized *horizontally* into square cells, but each cell stores a single
**continuous** depth interval per feature. Volumes are exact interval
arithmetic, not a count of occupied depth bins.

- **Panel A** — the study area. The Bangladesh EEZ on a Lambert Azimuthal Equal
  Area grid, each cell carrying its seafloor depth from GEBCO. An inset zooms in
  far enough to resolve individual square cells and highlights the one Panel B
  dissects.
- **Panel B** — that single cell's water column, drawn in isometric projection:
  the species depth interval, the fishery depth interval, and the intersection
  that `calc_volume_overlap()` returns.

The three quantities Panel B shows are exactly the ones the analysis computes:

| Panel B element | Analysis quantity |
|---|---|
| Purple slab | `depth_min` / `depth_max` from `voxelize_range()` for the species |
| Yellow slab | Same, for the fishery (after the 3 m tidal correction) |
| Teal slab | `depth_min_overlap` / `depth_max_overlap` from `calc_volume_overlap()` |
| Slab footprint | Cell area from `terra::cellSize()` |
| Blue floor | Seafloor depth, which truncates every slab from below |

## Setup

```{r setup}
# devtools::load_all()
library(sharkabc3d)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(patchwork)

sf::sf_use_s2(FALSE)
```

## Panel A: the study area grid

We rebuild the study grid the way the Bangladesh vignette does, so the schematic
describes the real analysis rather than an idealisation of it: the same GEBCO
release, the same EEZ polygon, and the same equal-area projection. The one
deliberate departure is the cell size — the analysis runs at 1000 m, but the
figure uses a coarser grid so that individual cells are visible in the inset.
That is a legibility choice, not a change to the method.

```{r study-grid}
bathy <- load_bathymetry(
  "~/Programming_Projects/Big_Data/gebco_2025_sub_ice_topo/GEBCO_2025_sub_ice.nc"
)

bangladesh_eez <- sf::st_read(
  "~/Programming_Projects/Big_Data/World_EEZ_v12_20231025/eez_v12.shp",
  query = "SELECT * FROM eez_v12 WHERE TERRITORY1 = 'Bangladesh'",
  quiet = TRUE
) %>% rename(geometry = "_ogr_geometry_")

# Lambert Azimuthal Equal Area centred on Bangladesh waters: this is what makes
# every cell the same size on the ground, and so what licenses treating
# the cell area as a constant when converting depth intervals to volumes.
bangladesh_crs <- "+proj=laea +lat_0=22 +lon_0=90 +datum=WGS84 +units=m"

# Grid resolution in metres. The Bangladesh analysis runs at 1000 m; this figure
# is free to use a coarser grid, because at 1 km the cells are far too small to
# resolve in the inset. Everything downstream — the highlighted cell footprint,
# the inset window, the cell area used for the volume arithmetic, and the panel
# labels — derives from this one number, so changing it here keeps the whole
# figure self-consistent.
grid_res <- 10000

study_grid <- create_study_raster(
  layers = list(bangladesh_eez),
  res = grid_res,
  crs = bangladesh_crs
)

# Crop before projecting so we never reproject the full global GEBCO grid.
# terra::project() resamples bilinearly by default, so each cell takes an
# interpolated seafloor depth rather than the mean of the ~450 m GEBCO cells it
# contains. clamp(lower = 0) puts land at zero depth.
bathy_crop <- terra::crop(bathy, terra::vect(bangladesh_eez))
seafloor <- terra::project(bathy_crop, study_grid)
seafloor <- terra::clamp(-seafloor, lower = 0)

# Mask to the EEZ: the analysis extent is the EEZ polygon, not the grid's
# bounding box, and the figure should show the same footprint.
eez_proj <- sf::st_transform(bangladesh_eez, bangladesh_crs)
seafloor_eez <- terra::mask(seafloor, terra::vect(eez_proj))
names(seafloor_eez) <- "depth"

# Read the realised cell size back off the raster rather than trusting
# `grid_res`: create_study_raster() fits whole cells to the layer extent, so the
# result can differ slightly from the requested resolution. These drive the
# highlighted cell, the inset window, and every "N km" label in the figure.
cell_dx <- terra::res(seafloor_eez)[1]
cell_dy <- terra::res(seafloor_eez)[2]

# Label for the cell size. Written as km when the cells are at least 1 km, and
# dropped to a whole number when it is one (so "10 km", not "10.0 km").
cell_label <- paste0(round(cell_dx / 1000, digits = 0), " km")
cell_area_km2 <- (cell_dx / 1000) * (cell_dy / 1000)

c(cell_dx = cell_dx, cell_dy = cell_dy, cell_area_km2 = cell_area_km2)
```

### Choosing the cell to dissect

Panel B needs one cell, and the two panels should agree: the seafloor depth
drawn in Panel B is read from the very cell the inset highlights. We pick a
mid-shelf cell — deep enough that a fishery does not reach the seabed, shallow
enough that the species' nominal lower limit gets truncated — because that is
the case worth explaining.

```{r focus-cell}
# A point on the mid shelf, in lon/lat, projected onto the study grid.
focus_lonlat <- sf::st_sfc(sf::st_point(c(90.5, 20.1)), crs = 4326)
focus_xy <- sf::st_coordinates(sf::st_transform(focus_lonlat, bangladesh_crs))

focus_cell <- terra::cellFromXY(seafloor_eez, focus_xy)
focus_centre <- terra::xyFromCell(seafloor_eez, focus_cell)
focus_depth <- terra::extract(seafloor_eez, focus_cell)[["depth"]]

# Round for the schematic: this is a conceptual figure, not a data readout.
focus_depth <- round(focus_depth / 10) * 10
focus_depth
#> [1] 450        # at grid_res = 10000; ~400 m at the analysis's 1000 m

# Footprint of the highlighted cell, as a polygon for both the inset and the
# leader line from Panel A to Panel B. Built from half a cell either side of the
# cell centre, so it tracks the grid resolution instead of assuming one.
focus_poly <- terra::as.polygons(
  terra::ext(
    focus_centre[1] - cell_dx / 2, focus_centre[1] + cell_dx / 2,
    focus_centre[2] - cell_dy / 2, focus_centre[2] + cell_dy / 2
  )
) |> sf::st_as_sf()
sf::st_crs(focus_poly) <- bangladesh_crs
```

### The full-extent map

At the full extent of the EEZ the cells are at or below pixel size, so we draw
the seafloor raster directly rather than thousands of cell polygons.
`coord_sf(datum = ...)` keeps the plot in the equal-area projection while
drawing a familiar lon/lat graticule.

```{r map-full}
# Box marking the region the inset magnifies. Sized in CELLS, not metres, so the
# inset always shows a legible number of them however the grid is resolved.
inset_cells <- 12
inset_ext <- terra::ext(
  focus_centre[1] - inset_cells / 2 * cell_dx,
  focus_centre[1] + inset_cells / 2 * cell_dx,
  focus_centre[2] - inset_cells / 2 * cell_dy,
  focus_centre[2] + inset_cells / 2 * cell_dy
)
inset_box <- sf::st_as_sf(terra::as.polygons(inset_ext))
sf::st_crs(inset_box) <- bangladesh_crs

depth_fill <- scale_fill_gradientn(
  name = "Seafloor\ndepth (m)",
  colours = c("#bcdcf5", "#4191d2", "#2166ac", "#0a2850"),
  na.value = "transparent",
  trans = "reverse"          # shallow pale at the top of the legend
)

p_map <- ggplot() +
  geom_sf(data = eez_proj, fill = "grey97", colour = "grey60",
          linewidth = 0.3, linetype = "dashed") +
  geom_spatraster(data = seafloor_eez) +
  depth_fill +
  geom_sf(data = inset_box, fill = NA, colour = "grey15", linewidth = 0.5) +
  geom_sf(data = focus_poly, fill = NA, colour = "#21918c", linewidth = 0.6) +
  coord_sf(datum = sf::st_crs(4326), expand = FALSE) +
  labs(title = paste0("A. Study area: ", cell_label, " equal-area grid")) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    axis.title = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "inside",
    legend.position.inside = c(0.82, 0.18),
    legend.background = element_rect(fill = "white", colour = "grey70"),
    legend.key.height = unit(0.9, "cm")
  )
p_map
```

### The inset: individual cells

Zoomed to a window a dozen cells across, the cells resolve. Here we *do* draw them as
polygons, with their boundaries visible, because the whole point of the inset is
that the grid is square and discrete — the single change from the earlier
hexagonal version of this analysis.

```{r map-inset}
inset_rast <- terra::crop(seafloor_eez, inset_ext)

# One polygon per cell, boundaries drawn. aggregate = FALSE keeps cells separate
# instead of dissolving equal-valued neighbours into one shape.
inset_cells <- terra::as.polygons(inset_rast, aggregate = FALSE, na.rm = TRUE) |>
  sf::st_as_sf()

p_inset <- ggplot() +
  geom_sf(data = inset_cells, aes(fill = depth),
          colour = "grey20", linewidth = 0.08) +
  depth_fill +
  geom_sf(data = focus_poly, fill = NA, colour = "#3ce8dc", linewidth = 1.1) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "none",
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8)
  )
p_inset
```

## Panel B: the water column of one cell

Panel B is an isometric drawing, built from ggplot polygons rather than a
rendered 3D scene. Two reasons: the slabs are translucent and deliberately
interpenetrating, which path tracers render unpredictably; and a polygon plot
exports as vector art, so the callout labels stay editable when the figure is
finished in Inkscape.

### Isometric projection helpers

An isometric view maps a 3D point to the page by shearing the two horizontal
axes apart and letting depth run straight up and down. Increasing `x` moves
right-and-back, increasing `y` moves left-and-back, and `z` (negative below the
sea surface) moves straight down. The viewer therefore sits over the
low-`x`, low-`y` corner and sees each box's top face plus its two near sides.

```{r iso-helpers}
ISO_C <- cos(pi / 6)   # horizontal shear of the x/y axes
ISO_S <- sin(pi / 6)   # vertical  shear of the x/y axes

# Page units per metre of depth. The cell footprint is 1 unit wide while the
# column is hundreds of metres deep, so depth has to be compressed hard or the
# drawing becomes a needle. Derived from the cell's actual depth rather than
# fixed, so the column keeps the same on-page height whatever bathymetry the
# focus cell turns out to have — this cell is ~450 m at 10 km resolution and
# ~400 m at the analysis's 1 km, and a hardcoded scale would misdraw both.
COLUMN_HEIGHT <- 3.2                     # target height of the column, in page units
VSCALE <- COLUMN_HEIGHT / focus_depth

iso <- function(x, y, z) {
  data.frame(
    px = (x - y) * ISO_C,
    py = (x + y) * ISO_S + z * VSCALE
  )
}

# Darken a colour toward black; used to shade the three visible faces of a box
# so it reads as a solid rather than a flat silhouette.
darken <- function(col, f) {
  grDevices::rgb(t(grDevices::col2rgb(col) * f), maxColorValue = 255)
}

# The three visible faces of an axis-aligned box, as a polygon data frame.
# z_top and z_bot are depths in metres below the surface (positive down); the
# box spans from z_top down to z_bot.
iso_box <- function(x0, x1, y0, y1, z_top, z_bot, colour) {
  zt <- -z_top
  zb <- -z_bot

  face <- function(pts, shade, id) {
    d <- iso(pts[, 1], pts[, 2], pts[, 3])
    d$grp <- id
    d$fill <- darken(colour, shade)
    d
  }

  rbind(
    # Top face: brightest.
    face(cbind(c(x0, x1, x1, x0), c(y0, y0, y1, y1), rep(zt, 4)), 1.00, "top"),
    # y = y0 face, falling away to the right of the viewer.
    face(cbind(c(x0, x1, x1, x0), rep(y0, 4), c(zb, zb, zt, zt)), 0.82, "right"),
    # x = x0 face, falling away to the left of the viewer.
    face(cbind(rep(x0, 4), c(y0, y1, y1, y0), c(zb, zb, zt, zt)), 0.64, "left")
  )
}

# Draw one box as three shaded polygons at a fixed transparency.
geom_iso_box <- function(box, alpha = 0.55) {
  geom_polygon(
    data = box,
    aes(x = px, y = py, group = grp, fill = fill),
    alpha = alpha, colour = "grey25", linewidth = 0.2
  )
}
```

### The synthetic depth ranges

The depths below are invented for legibility; only the seafloor depth is real,
read from the highlighted cell in Panel A. They are chosen to show the two
behaviours that matter in the analysis:

1. The species' nominal lower limit is compared against the seafloor, and
   `voxelize_range()` truncates it at the bed when it is deeper. Whether that
   bites depends on the cell: this one is ~450 m, so a 250 m limit is **not**
   clamped. Raise `species_bot_nominal` past the seafloor depth to show the
   truncation, which is the behaviour the panel is meant to illustrate.
2. The fishery works the full water column down to the bed, so it encloses the
   species' whole depth range. The intersection is therefore the entire species
   interval: this cell offers **no depth refuge**, and the species is fully
   exposed to the gear where the two footprints meet.

```{r depth-ranges}
seafloor_m <- focus_depth        # from Panel A's highlighted cell (450 m here)

species_top <- 80
species_bot_nominal <- 250                            # IUCN / study depth limit
species_bot <- min(species_bot_nominal, seafloor_m)   # clamped by voxelize_range()

fishery_top <- 0
fishery_bot <- 120

# The intersection calc_volume_overlap() computes: interval arithmetic on the
# two already-clamped intervals.
overlap_top <- max(species_top, fishery_top)
overlap_bot <- min(species_bot, fishery_bot)
overlap_thickness <- max(0, overlap_bot - overlap_top)

c(species = species_bot - species_top,
  fishery = fishery_bot - fishery_top,
  overlap = overlap_thickness)
#>  species  fishery  overlap
#>      170      120       40

# Volumes, in the km^3 the analysis reports: the cell's footprint area times the
# interval thickness expressed in km. cell_area_km2 comes from the study grid
# above, so these track the chosen resolution.
c(species_km3 = cell_area_km2 * (species_bot - species_top) / 1000,
  overlap_km3 = cell_area_km2 * overlap_thickness / 1000)
```

### Assembling the scene

The three slabs share the cell's footprint but are offset horizontally so
all three stay visible — an exploded view, not a claim that they occupy
different ground. The overlap slab is drawn last and least transparent, since it
is the quantity the figure exists to explain.

```{r iso-scene}
col_species   <- "#7b52ab"   # purple, matching the vignette's overlap maps
col_fishery   <- "#e8d54a"   # yellow
col_intersect <- "#21918c"   # teal
col_water     <- "#7fb2e5"
col_bed       <- "#3b5aa0"

# How far the species and fishery slabs are pulled apart. The offsets must be
# OPPOSED in x and y: this projection has px = (x - y) * ISO_C, so shifting both
# axes by the same amount moves a box straight up the page and separates nothing.
# Shifting -x/+y sends the species left, +x/-y sends the fishery right, and both
# stay at the same height because py depends on (x + y), which is unchanged.
#
# At off = 0 all three slabs share one footprint exactly, which is geometrically
# truthful but hides the intersection inside the other two. Raise this to pull
# the species left and the fishery right and expose the overlap slab between.
off <- 0

box_species <- iso_box(-off, 1 - off,  off, 1 + off,
                       species_top, species_bot, col_species)
box_fishery <- iso_box( off, 1 + off, -off, 1 - off,
                       fishery_top, fishery_bot, col_fishery)
box_overlap <- iso_box(0, 1, 0, 1,
                       overlap_top, overlap_bot, col_intersect)

# The sea surface: a translucent plane at z = 0. Its projected half-width is
# surf * 2 * ISO_C, so keep it modest or the plane dwarfs the column.
surf <- 1.5
water_plane <- iso(c(-surf, surf, surf, -surf),
                   c(-surf, -surf, surf, surf), rep(0, 4))

# The cell's seafloor: a thin plate at the bed, capping the column from below.
bed_plate <- iso_box(0, 1, 0, 1, seafloor_m, seafloor_m + 4, col_bed)

# The cell footprint at the surface, and guide lines dropping from its corners
# to the bed, so the column the slabs belong to is unambiguous.
surface_cell <- iso(c(0, 1, 1, 0), c(0, 0, 1, 1), rep(0, 4))

corners <- expand.grid(x = c(0, 1), y = c(0, 1))
guides <- do.call(rbind, lapply(seq_len(nrow(corners)), function(i) {
  d <- iso(rep(corners$x[i], 2), rep(corners$y[i], 2), c(0, -seafloor_m))
  d$grp <- i
  d
}))
```

### Callouts

Each element gets a label anchored off to one side with a leader line running to
the feature it names, as in the manuscript figure. There is no separate depth
axis: the depths are carried in the label text instead, which keeps the drawing
uncluttered and avoids an axis colliding with the callouts.

Because the projection maps a 3D point to the page, target points are computed
with `iso()` rather than guessed — that way they stay attached to the geometry if
the depths or offsets above are changed. Label anchors are still tuned by eye.

```{r iso-annotations}
# Helper: a point on the geometry, in page coordinates.
at <- function(x, y, depth) iso(x, y, -depth)

# Mid-edge of each slab's visible near face, used as leader targets: the species
# slab is read off its left face (x = -off), the fishery off its right face
# (y = -off).
sp_left_x <- -off;      sp_mid_y <-  off + 0.5
fi_right_y <- -off;     fi_mid_x <-  off + 0.5

targets <- rbind(
  surface   = at(-surf * 0.8, surf * 0.8, 0),          # far corner of the water plane
  sp_top    = at(sp_left_x, sp_mid_y, species_top),
  sp_bot    = at(sp_left_x, sp_mid_y, species_bot),
  fi_top    = at(fi_mid_x, fi_right_y, fishery_top),
  fi_bot    = at(fi_mid_x, fi_right_y, fishery_bot),
  overlap   = at(0, 0, (overlap_top + overlap_bot) / 2),
  bed       = at(0, 0, seafloor_m),
  cell      = at(1, 0, 0)
)

# Label anchors: left column hjust = 1, right column hjust = 0.
label_x_left  <- -2.6
label_x_right <-  2.6

callouts <- data.frame(
  key = rownames(targets),
  label = c(
    "Sea surface (0 m)",
    paste0("Upper depth of species (", species_top, " m)"),
    # Only claim clamping when it actually happened: if the nominal limit is
    # shallower than this cell's bed, voxelize_range() leaves it alone and the
    # label would otherwise assert a truncation that never took place.
    if (species_bot_nominal > seafloor_m) {
      paste0("Lower depth of species\nclamped to seafloor (", species_bot, " m)")
    } else {
      paste0("Lower depth of species (", species_bot, " m)")
    },
    paste0("Upper depth of fishery (", fishery_top, " m)"),
    paste0("Lower depth of fishery (", fishery_bot, " m)"),
    paste0("Overlap volume (", overlap_top, "-", overlap_bot, " m)"),
    paste0("Seafloor depth of grid cell (", seafloor_m, " m)"),
    paste0(cell_label, " grid cell")
  ),
  lx = c(label_x_left, label_x_left, label_x_left,
         label_x_right, label_x_right,
         label_x_left, label_x_right, label_x_right),
  ly = c(0.95, -0.75, -2.60,
         0.35, -1.05,
         -1.70, -2.90, 1.20),
  tx = targets$px,
  ty = targets$py
)
callouts$hjust <- ifelse(callouts$lx < 0, 1, 0)
```

```{r panel-b}
p_column <- ggplot() +
  # Water surface, drawn first so everything else sits over it.
  geom_polygon(data = water_plane, aes(px, py),
               fill = col_water, alpha = 0.28, colour = NA) +
  # Column guide lines from the surface cell corners down to the bed.
  geom_path(data = guides, aes(px, py, group = grp),
            colour = "grey45", linewidth = 0.25, linetype = "22") +
  # Seafloor plate.
  geom_iso_box(bed_plate, alpha = 0.95) +
  # Fishery is furthest from the viewer, so it goes down first.
  geom_iso_box(box_fishery, alpha = 0.45) +
  geom_iso_box(box_species, alpha = 0.45) +
  # The intersection last and most opaque: it is the point of the figure.
  geom_iso_box(box_overlap, alpha = 0.80) +
  # Cell footprint at the surface.
  # geom_polygon(data = surface_cell, aes(px, py),
  #              fill = "#3ce8dc", alpha = 0.75,
  #              colour = "#0f6b66", linewidth = 0.5) +
  scale_fill_identity() +
  # Callout leader lines, then the labels over them. geom_label (rather than
  # annotate) so label.size = NA is honoured and the boxes stay borderless.
  geom_segment(data = callouts,
               aes(x = lx, xend = tx, y = ly, yend = ty),
               colour = "grey30", linewidth = 0.3) +
  geom_label(data = callouts,
             aes(x = lx, y = ly, label = label, hjust = hjust),
             size = 3, label.size = NA, fill = "white", alpha = 0.85,
             lineheight = 0.95) +
  coord_equal(clip = "off") +
  labs(title = "B. Depth overlap within one grid cell") +
  theme_void(base_size = 11) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 12),
    plot.margin = margin(10, 90, 10, 90)
  )
p_column
```

## Composing the figure

The map carries the inset in its upper-right corner via
`patchwork::inset_element()`, then sits beside the isometric panel. The two get
roughly equal width, as in the manuscript layout.

```{r compose-figure, fig.width = 14, fig.height = 8}
p_map_inset <- p_map +
  inset_element(p_inset, left = 0.52, bottom = 0.62, right = 0.99, top = 0.99)

fig_overlap_schematic <- (p_map_inset | p_column) +
  plot_layout(widths = c(1, 1.05))
fig_overlap_schematic

# Always write into the figures/ directory, whether render runs from the project
# root or from figures/ itself.
fig_dir <- if (basename(getwd()) == "figures") "." else "figures"
ggsave(
  filename = file.path(fig_dir, "fig-overlap-schematic_raw.pdf"),
  plot = fig_overlap_schematic,
  width = 14, height = 8, dpi = 300, bg = "white"
)
ggsave(
  filename = file.path(fig_dir, "fig-overlap-schematic_raw.png"),
  plot = fig_overlap_schematic,
  width = 14, height = 8, dpi = 300, bg = "white"
)
```

`fig-overlap-schematic_raw.pdf` is the raw export; hand-edit it in Inkscape to
produce the final `fig-overlap-schematic.pdf` / `.png` used in the manuscript.
The callout label positions in `iso-annotations` are tuned by eye and are the
most likely thing to need nudging there.

## Recap

- The grid is **square** and equal-area — the change from the earlier
  hexagonal version of this analysis.
- Each cell holds one **continuous** depth interval per feature. Nothing is
  snapped to standard depth levels; the vertical dimension is analytical, and
  only the horizontal is discretized. That is what makes the model 2.5D rather
  than a true voxel stack.
- `voxelize_range()` truncates every interval at the cell's seafloor before any
  overlap is computed. Whether that truncation is visible in Panel B depends on
  the focus cell: it bites only when the nominal limit is deeper than that
  cell's bed.
- `calc_volume_overlap()` intersects the two clamped intervals and multiplies
  the resulting thickness by the cell area. Summed over cells, that is the
  overlap volume the analysis reports.
- The gap between the teal slab and the seafloor is the **depth refuge**: water
  the species occupies that the fishery does not reach.
