---
title: "Voxelizing a Species Range"
output: html_document
---

<!--
This is NOT a package vignette. It is the source that builds Figure 2 of the
accompanying manuscript (figures/fig-species-voxel.png). It lives in figures/ and
is excluded from the built package via .Rbuildignore. Render it to regenerate the
figure (ggsave writes into figures/ whether run from the project root or here); it
depends on rayrender, patchwork, and sf. It reuses the same synthetic study area
(grid, bathymetry, standard depths) as figures/voxel-model-visualization.Rmd
(Figure 1), so the two figures share one worked example.
-->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Voxelizing a Species Range

Figure 1 built the **voxel model** of the ocean: a stack of raster layers, one
per standard depth, clamped to the seafloor. This figure shows how the package
turns a **species** into that same representation — a *voxelized range*. Three
inputs combine:

1. a **range polygon** — where the species is found, horizontally (Panel A);
2. a **depth range** — the band of the water column it occupies (Panel B); and
3. the **voxel ocean model** from Figure 1 — the discretised study space that
   caps every column at the seafloor (Panel C).

`voxelize_range()` intersects the three: a voxel is *present* when its cell lies
inside the polygon **and** its standard depth falls within the species' depth
band **and** the water column there reaches that depth. The result (Panel D) is
a raster stack whose cells are coloured where the species is present and left
empty (NA) where it is not — the object `calc_volume()` and
`calc_volume_overlap()` measure.

Both 3D panels use [rayrender](https://www.rayrender.net/), rendering each voxel
as a solid plate exactly as in Figure 1.

## Setup

```{r setup}
# devtools::load_all()
library(sharkabc3d)
library(terra)
library(sf)
library(ggplot2)
library(patchwork)
library(rayrender)
library(png)
```

## The study area (shared with Figure 1)

We rebuild the same synthetic 10 x 10 study area as Figure 1: an empty grid
template, a GEBCO-style bathymetry that deepens to ~400 m in the south-east with
a small landmass in the north-west, and ten standard depths. See
`voxel-model-visualization.Rmd` for the reasoning behind each input.

```{r study-area}
template <- terra::rast(
  nrow = 10, ncol = 10,
  xmin = 0, xmax = 10, ymin = 0, ymax = 10,
  crs = "EPSG:4326"
)

xy <- terra::xyFromCell(template, seq_len(terra::ncell(template)))
x  <- xy[, 1]
y  <- xy[, 2]
elev <- -(30 + (x^1.7) * 5 + (10 - y) * 15)   # seafloor deepens to the south-east
elev[x < 2.5 & y > 7.5] <- 20                 # small landmass in the north-west
elev <- round(elev / 25) * 25                 # snap to 25 m terraces

bathymetry <- terra::setValues(terra::rast(template), elev)
names(bathymetry) <- "elevation"

depths <- c(0, 25, 50, 100, 150, 200, 250, 300, 350, 400)

voxel <- create_study_voxel(
  template   = template,
  bathymetry = bathymetry,
  depths     = depths
)
```

## The species: a range polygon and a depth band

The species' horizontal range is an irregular oval over the grid (the same noisy
ellipse used elsewhere, with a fixed seed for reproducibility). Its depth range
is a band of the water column — here 50 m to 250 m, a mid-water species that
avoids the surface and does not reach the deep south-east floor.

```{r species}
# Horizontal range: an irregular oval centred on the grid.
set.seed(42)
n_vertices <- 24
ang    <- seq(0, 2 * pi, length.out = n_vertices + 1)[-(n_vertices + 1)]
noise  <- runif(n_vertices, 0.75, 1.15)
poly_pts <- cbind(
  x = 5 + 3.2 * noise * cos(ang),
  y = 5 + 2.4 * noise * sin(ang)
)
poly_pts <- rbind(poly_pts, poly_pts[1, ])   # close the ring

sp_poly <- sf::st_sf(
  id = 1L,
  geometry = sf::st_sfc(sf::st_polygon(list(poly_pts)), crs = "EPSG:4326")
)

# Vertical range: the species' depth band, in metres.
depth_min <- 50
depth_max <- 250

# A warm highlight colour for the species (contrasts the ocean's blues).
species_col <- "#e6550d"
```

## Visualizing the inputs (Panels A and B)

A shared, muted theme matches Figure 1.

```{r input-theme}
# Match the hand-edited style of Figure 1: Liberation Sans (the Arial-metric font
# the final PDF embeds) and teal panel titles. Rendering via cairo_pdf (see the
# compose chunk) embeds the system font, so the raw export already looks close to
# the finished figure.
title_font <- "Liberation Sans"
teal_col   <- "#1a9e9e"

theme_voxel_input <- function() {
  theme_minimal(base_size = 11, base_family = title_font) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(family = title_font, colour = teal_col, size = 12)
    )
}
```

### Panel A: the species range polygon

The polygon over the grid template: where the species is found, horizontally.

```{r plot-range}
grid_cells <- terra::as.polygons(template, aggregate = FALSE) |>
  sf::st_as_sf()

p_range <- ggplot() +
  geom_sf(data = grid_cells, fill = "grey97", colour = "grey85",
          linewidth = 0.2) +
  geom_sf(data = sp_poly, fill = scales::alpha(species_col, 0.30),
          colour = species_col, linewidth = 0.7) +
  coord_sf(expand = FALSE) +
  labs(title = "A. Species range") +
  theme_voxel_input() +
  theme(axis.title = element_blank())
p_range
```

### Panel B: the species depth range

An inverted depth axis (surface at the top) with the standard-depth levels
labelled on the axis, the whole water column shown faintly, and the species'
50–250 m band highlighted.

```{r plot-depthrange}
p_depthrange <- ggplot() +
  # Full water column as a faint bar, flush against the depth axis (xmin = 0,
  # no left expansion) so there is no gap between the bar and the y axis.
  annotate("rect", xmin = 0, xmax = 0.35, ymin = 0, ymax = max(depths),
           fill = "grey92") +
  # The species' depth band.
  annotate("rect", xmin = 0, xmax = 0.35, ymin = depth_min, ymax = depth_max,
           fill = species_col, alpha = 0.9) +
  annotate("text", x = 0.45, y = (depth_min + depth_max) / 2,
           label = paste0(depth_min, "–", depth_max, " m"),
           hjust = 0, size = 3.2, colour = species_col, fontface = "bold",
           family = title_font) +
  # Break the depth axis at the standard depths so each layer is labelled.
  scale_y_reverse(breaks = depths, limits = c(max(depths), -10)) +
  scale_x_continuous(limits = c(0, 1.3), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "B. Species depth range", y = "Depth (m)") +
  theme_voxel_input() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())
p_depthrange
```

## Expanding the voxel model and masking to the species

We expand the `study_voxel` into one row per (ocean cell x standard-depth
layer), exactly as in Figure 1, then flag which voxels the species occupies: the
cell must fall inside the range polygon and the depth must lie within the band.
Reachability (depth <= seafloor) is already guaranteed by the expansion.

```{r voxel-columns}
depth_bands <- function(standard_depths) {
  sd <- sort(unique(standard_depths))
  n  <- length(sd)
  mids <- (sd[-n] + sd[-1]) / 2
  data.frame(
    depth  = sd,
    z_low  = c(0, mids),
    z_high = c(mids, sd[n] + (sd[n] - sd[n - 1]) / 2)
  )
}

voxel_columns <- function(voxel) {
  bands <- depth_bands(voxel$depths)
  cells <- terra::as.data.frame(voxel$seafloor, xy = TRUE, na.rm = FALSE)
  do.call(rbind, lapply(seq_len(nrow(cells)), function(i) {
    sfd <- cells$seafloor[i]
    if (is.na(sfd) || sfd <= 0) return(NULL)      # skip land / dry cells
    keep <- bands$z_low < sfd
    if (!any(keep)) return(NULL)
    b <- bands[keep, , drop = FALSE]
    b$z_high <- pmin(b$z_high, sfd)               # clamp deepest voxel to seafloor
    data.frame(x = cells$x[i], y = cells$y[i],
               depth = b$depth, z_low = b$z_low, z_high = b$z_high)
  }))
}

vox <- voxel_columns(voxel)

# Which voxels the species occupies: cell inside the polygon AND depth in band.
sp_mask   <- terra::rasterize(terra::vect(sp_poly), template, field = 1)
in_poly   <- !is.na(terra::extract(sp_mask, cbind(vox$x, vox$y))[, 1])
vox$present <- in_poly & vox$depth >= depth_min & vox$depth <= depth_max

c(total_voxels = nrow(vox), present_voxels = sum(vox$present))
#> total_voxels present_voxels
#>          558             83
```

## Rendering the 3D panels (C and D)

The rayrender machinery is identical to Figure 1: one solid plate per voxel,
placed at its true depth, viewed down the same camera angles so the panels line
up. Both panels wireframe the bounding box of the study extent so the voxels read
as cells inside a defined volume: Panel C colours every ocean voxel by depth (the
voxel model), while Panel D colours only the *present* voxels in the species
highlight colour and leaves the rest empty.

```{r voxel-scene}
# Depth -> colour for the ocean model (Panel C): shallow pale blue, deep navy.
depth_palette <- grDevices::colorRampPalette(
  c("#bcdcf5", "#3a7fc0", "#0a2f66")
)(256)
depth_to_colour <- function(d, dmin, dmax) {
  idx <- pmax(1, pmin(256, ceiling((d - dmin) / (dmax - dmin) * 256)))
  depth_palette[idx]
}

res     <- terra::res(voxel$grid)
dx      <- res[1]
dy      <- res[2]
vscale  <- 6 / max(vox$z_high)          # true depths -> ~6 scene units at the seafloor
hgap    <- 0.94                         # horizontal seams so raster cells stay visible
plate_h <- 0.07                         # thickness of each solid layer plate
dmin    <- min(voxel$depths)
dmax    <- max(voxel$depths)

# One solid plate per voxel at its true depth (clamped to the seafloor), in a
# given colour. Shared by both scenes.
voxel_plate <- function(r, colour) {
  rayrender::cube(
    x = r$x, y = -pmin(r$depth, r$z_high) * vscale, z = r$y,
    xwidth = dx * hgap, ywidth = plate_h, zwidth = dy * hgap,
    material = rayrender::diffuse(color = colour)
  )
}

# Panel C: the full voxel ocean, every plate coloured by depth.
ocean_cubes <- lapply(seq_len(nrow(vox)), function(i) {
  r <- vox[i, ]
  voxel_plate(r, depth_to_colour(r$depth, dmin, dmax))
})

# Panel D: only the voxels the species occupies, in the highlight colour. Absent
# voxels are simply not drawn, so they read as empty space.
present_vox    <- vox[vox$present, ]
species_cubes  <- lapply(seq_len(nrow(present_vox)), function(i) {
  voxel_plate(present_vox[i, ], species_col)
})

# Wireframe the bounding box of the study extent, so the voxels read as cells
# inside a defined volume rather than shapes floating in space. Just a rectangle
# at the surface and at the deepest standard depth, joined by four vertical corner
# edges (thin cylinders via rayrender::segment). Shared by both 3D panels.
ext      <- as.vector(terra::ext(voxel$grid))   # xmin, xmax, ymin(=z), ymax(=z)
y_bottom <- -max(depths) * vscale
wire_col <- "#404040"   # dark grey: clearly visible against the white background
wire_r   <- 0.012
wire_seg <- function(a, b) {
  rayrender::segment(start = a, end = b, radius = wire_r,
                     material = rayrender::diffuse(color = wire_col))
}
rect_at <- function(yd) list(
  wire_seg(c(ext[1], yd, ext[3]), c(ext[2], yd, ext[3])),
  wire_seg(c(ext[2], yd, ext[3]), c(ext[2], yd, ext[4])),
  wire_seg(c(ext[2], yd, ext[4]), c(ext[1], yd, ext[4])),
  wire_seg(c(ext[1], yd, ext[4]), c(ext[1], yd, ext[3]))
)
corners <- list(c(ext[1], ext[3]), c(ext[2], ext[3]),
                c(ext[2], ext[4]), c(ext[1], ext[4]))
wire_objs <- c(
  rect_at(0),          # top face at the surface
  rect_at(y_bottom),   # bottom face at the deepest standard depth
  lapply(corners, function(cc) wire_seg(c(cc[1], 0, cc[2]),
                                        c(cc[1], y_bottom, cc[2])))
)
wireframe <- Reduce(rayrender::add_object, wire_objs)

# A soft overhead light, shared by both scenes.
key_light <- rayrender::sphere(
  x = 14, y = 20, z = -10, radius = 3,
  material = rayrender::light(intensity = 8)
)

scene_ocean   <- Reduce(rayrender::add_object, ocean_cubes)   |>
  rayrender::add_object(wireframe) |>
  rayrender::add_object(key_light)
scene_species <- Reduce(rayrender::add_object, species_cubes) |>
  rayrender::add_object(wireframe) |>
  rayrender::add_object(key_light)
```

```{r voxel-render}
# Camera angles shared across the 3D panels (as in Figure 1): compass azimuth and
# a shallow elevation so the eye sits low and near side-on, opening up the layers.
view_theta <- 35
view_phi   <- 18
rad     <- pi / 180
cam_dir <- c(
   cos(view_phi * rad) * sin(view_theta * rad),   # east
   sin(view_phi * rad),                           # up
  -cos(view_phi * rad) * cos(view_theta * rad)    # north (rgl south -> negate)
)

# rayrender comes out mirrored east-west relative to rayshader's convention, so
# we flip the finished image left-right (see Figure 1).
flip_h <- function(img) img[, rev(seq_len(dim(img)[2])), , drop = FALSE]

render_voxels <- function(scene, file, lookat_y, dist, fov, width, height) {
  rayrender::render_scene(
    scene,
    lookfrom       = c(5, 0, 5) + dist * cam_dir,
    lookat         = c(5, lookat_y, 5),
    fov            = fov,
    samples        = 256,
    width          = width, height = height,
    ambient_light  = TRUE,
    backgroundhigh = "#ffffff", backgroundlow = "#ffffff",
    filename       = file
  )
  flip_h(png::readPNG(file))
}

ocean_png   <- tempfile(fileext = ".png")
species_png <- tempfile(fileext = ".png")

# Both panels share one camera/frame (as D and E do in Figure 1) so the 10x10
# grid footprint reads at the same size and the species subset is comparable to
# the full ocean.
img_ocean   <- render_voxels(scene_ocean, ocean_png,
                             lookat_y = -3, dist = 42, fov = 30,
                             width = 880, height = 1200)
img_species <- render_voxels(scene_species, species_png,
                             lookat_y = -3, dist = 42, fov = 30,
                             width = 880, height = 1200)

p_ocean <- patchwork::wrap_elements(
  full = grid::rasterGrob(img_ocean, interpolate = TRUE)) +
  labs(title = "C. Voxelized ocean model") +
  theme(plot.title.position = "plot",
        plot.title = element_text(family = title_font, colour = teal_col, size = 12))

p_species <- patchwork::wrap_elements(
  full = grid::rasterGrob(img_species, interpolate = TRUE)) +
  labs(title = "D. Voxelized species range") +
  # D is the output, not an input, so its title is black rather than teal.
  theme(plot.title.position = "plot",
        plot.title = element_text(family = title_font, colour = "black", size = 12))
```

Panel D is the intersection of the three inputs: the species' plates sit inside
the polygon's footprint (Panel A), span only the 50–250 m band (Panel B), and are
clamped to the seafloor of the voxel ocean (Panel C). The light bounding-box
wireframe (shared with Panel C) marks the study extent, so the coloured voxels
read as a subset of a defined volume; everywhere else the stack is empty, the NA
cells the package skips when it sums presence into a volume.

## Composing the manuscript figure

Two columns: the three inputs (A range polygon, B depth range, C voxel ocean
model) stacked in a narrow left column, and the voxelized species range (D) as a
large right column at double the width. The figure is sized so the right column's
aspect matches D's portrait render, so D fills the column. Mirroring the
hand-edited Figure 1, the inputs are wrapped in a teal-bordered box labelled
"INPUTS".

```{r compose-figure, fig.width = 12, fig.height = 11}
# Group the three inputs in a teal-bordered box titled "INPUTS": plot_annotation
# draws the border (plot.background) and the title, and wrap_elements turns the
# annotated group into a single patch so it sits beside the large output panel D.
inputs_group <- (p_range / p_depthrange / p_ocean) +
  plot_annotation(
    title = "INPUTS",
    theme = theme(
      plot.title = element_text(family = title_font, colour = teal_col,
                                face = "bold", hjust = 0.5, size = 14),
      plot.background = element_rect(colour = teal_col, fill = NA, linewidth = 1),
      plot.margin = margin(6, 8, 6, 8)
    )
  )

fig_species <- wrap_elements(full = inputs_group) | p_species
fig_species <- fig_species +
  plot_layout(widths = c(1, 2))
fig_species

# Always write into the figures/ directory, whether render runs from the project
# root or from figures/ itself. Render with cairo_pdf so the system font
# (Liberation Sans) embeds, matching the final figure.
fig_dir <- if (basename(getwd()) == "figures") "." else "figures"
ggsave(
  filename = file.path(fig_dir, "fig-species-voxel_raw.pdf"),
  plot = fig_species,
  width = 12, height = 11, dpi = 300, bg = "white",
  device = cairo_pdf
)

# Figures modified manually in vector editing software (Inkscape) after saving.
```

`fig-species-voxel_raw.pdf` is the raw export; hand-edit it in Inkscape to produce
the final `fig-species-voxel.pdf` / `.png` used in the manuscript.

## Recap

- A voxelized range is the voxel model masked to a species: cells inside the
  **range polygon**, at depths within the species' **depth band**, clamped to the
  seafloor of the **voxel ocean**.
- [voxelize_range()] performs that intersection; present cells carry the species,
  absent cells are NA (empty space).
- The resulting stack is what [calc_volume()] and [calc_volume_overlap()] measure
  — the same representation as the environmental stacks in Figure 1.
```
