---
title: "Visualizing the Voxel Spatial Model"
output: html_document
---

<!--
This is NOT a package vignette. It is the source that builds Figure 1 of the
accompanying manuscript (figures/fig-voxel-model.png). It lives in figures/ and
is excluded from the built package via .Rbuildignore. Render it (from this
directory, so ggsave writes the figure here) to regenerate the figure; it
depends on rayshader, rayrender, tidyterra, patchwork, and sf.
-->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Visualizing the Voxel Spatial Model

The spatial representation at the core of `sharkabc3d` is the **voxel model**:
a cube representation of three-dimensional space, the volumetric analogue of
the cell (or pixel) in a two-dimensional raster. Each cube is a *voxel*.

Every 3D operation in the package is built on this model, and every model is
built by [create_study_voxel()], which bundles **three inputs** into one
`study_voxel` object:

1. a **horizontal grid template** — the map footprint each voxel column sits on;
2. a **bathymetry raster** — the seafloor depth that caps each column; and
3. a vector of **standard depths** — the vertical levels that slice the water
   column into layers.

This document builds a deliberately small, synthetic example so the geometry
is easy to see, visualizes each of the three inputs on its own, then renders
the **output**: the voxelized ocean that the package computes volumes and
overlaps on. It reproduces Figure 1 of the accompanying manuscript.

The 3D panels use the [rayshader](https://www.rayshader.com/) ecosystem:
`rayshader::plot_3d()` renders the bathymetry input as a shaded surface (its
native strength), and [rayrender](https://www.rayrender.net/) — rayshader's
path-tracing sister package by the same author — builds the voxel output as
true stacked cubes. Both render straight to a PNG file, which we read back and
tile with the 2D panels using `patchwork`.

## Setup

```{r setup}
# devtools::load_all()
library(sharkabc3d)
library(terra)
library(ggplot2)
library(tidyterra)
library(patchwork)
library(rayshader)
library(rayrender)
```

## A simplified study area

Rather than load a multi-gigabyte GEBCO grid, we hand-build a 10 x 10 study
area. The extent, resolution, and coordinate system are arbitrary here — the
point is to keep the cube count low enough to read at a glance. A real analysis
would supply the same three inputs at full resolution (see the Bangladesh,
World Ocean Atlas, and Global Fishing Watch vignettes).

### Input 1: the horizontal grid template

The template is an empty raster. It defines the extent, resolution, and CRS
whose cells become the horizontal footprint of each voxel column. In a real
workflow it comes from [create_study_raster()], which sizes a grid to cover
your input layers; here we declare one directly.

```{r template}
template <- terra::rast(
  nrow = 10, ncol = 10,
  xmin = 0, xmax = 10, ymin = 0, ymax = 10,
  crs = "EPSG:4326"
)
template
```

### Input 2: the bathymetry

Bathymetry gives the depth of the seafloor at each cell, and so sets how tall
each voxel column can be. We synthesize a GEBCO-style elevation raster
(negative values below sea level, positive on land): a shallow shelf in the
north-west that drops down a slope to roughly 400 m in the south-east, plus a
small landmass in the north-west corner. We then snap the depths to 25 m steps,
so neighbouring cells of equal depth form flat terraces — the discrete,
per-cell depths the voxel model actually works on, where each cell's value is
its distance below the surface.

```{r bathymetry}
xy <- terra::xyFromCell(template, seq_len(terra::ncell(template)))
x  <- xy[, 1]
y  <- xy[, 2]

# Seafloor deepens toward the south-east: a shallow shelf drops down a slope to
# ~400 m in the far corner. Values are metres relative to sea level.
elev <- -(30 + (x^1.7) * 5 + (10 - y) * 15)

# A small landmass in the north-west corner (positive elevation = land).
elev[x < 2.5 & y > 7.5] <- 20

# Snap to 25 m steps: equal-depth cells form flat terraces, so the render reads
# as a staircase whose step heights are depths below the surface.
elev <- round(elev / 25) * 25

bathymetry <- terra::setValues(terra::rast(template), elev)
names(bathymetry) <- "elevation"
bathymetry
```

### Input 3: the standard depths

The standard depths are the vertical levels that define the model's vertical
resolution — the same idea as the standard depths of an oceanographic product
such as the World Ocean Atlas. Here we use ten levels: a slightly finer 25 m
spacing near the surface, where conditions change fastest, then a regular 50 m
spacing through the deeper water so the levels are easy to read.

```{r depths}
depths <- c(0, 25, 50, 100, 150, 200, 250, 300, 350, 400)
depths
```

## Visualizing the three inputs

Before assembling the voxel model, we plot each input on its own so the roles
are clear. A shared, muted theme keeps the panels legible when tiled together.

```{r input-theme}
theme_voxel_input <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      # Align the title to the plot's left edge (not the panel), so the A/B/C
      # headings line up with each other and with the plots' left side.
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = 12)
    )
}
```

### Panel A: horizontal grid template

The template carries no values — only a grid. We draw the cell boundaries to
convey the horizontal footprint that each voxel column will stand on.

```{r plot-template}
grid_cells <- terra::as.polygons(template, aggregate = FALSE) |>
  sf::st_as_sf()

p_template <- ggplot() +
  geom_sf(data = grid_cells, fill = "grey97", colour = "grey55",
          linewidth = 0.25) +
  coord_sf(expand = FALSE) +
  labs(title = "A. Horizontal grid template") +
  theme_voxel_input() +
  theme(axis.title = element_blank())
p_template
```

### Panel B: bathymetry (rayshader surface)

Bathymetry is a surface, so we render it as one with `rayshader` — but we want
the render to show the model's *rasterised* nature, not a smoothly interpolated
seafloor. By default `plot_3d()` interpolates between cell centres; upsampling
the raster with **nearest-neighbour** resampling first copies each cell value
across a block, so the only height changes happen at cell boundaries. The render
then draws flat steps joined by near-vertical walls: a staircase whose every
step sits at that cell's depth below the surface. A translucent water plane at
sea level (`water = TRUE, waterdepth = 0`) shows the ocean sitting over the
stepped seafloor — the surface that caps each voxel column from below. The
render is written straight to a PNG we reuse in the composite.

```{r plot-bathymetry}
# Render each cell as a flat block rather than a smooth surface: nearest-
# neighbour upsampling copies each cell value across a 12 x 12 patch, so heights
# change only at cell boundaries and plot_3d draws flat steps joined by near-
# vertical walls (the rasterised staircase).
bathymetry_blocky <- terra::disagg(bathymetry, fact = 12, method = "near")

# rayshader reads a plain elevation matrix (no CRS): negative below sea level,
# positive on land, exactly the GEBCO convention this raster already uses.
elev_mat <- rayshader::raster_to_matrix(bathymetry_blocky)

# Colour ramp keyed to elevation: deep = dark blue, shallow = pale, land = tan.
bathy_texture <- grDevices::colorRampPalette(
  c("#0a2850", "#2166ac", "#4191d2", "#c6e6ff", "#d9c8a0")
)(256)
bathy_shade <- rayshader::height_shade(elev_mat, texture = bathy_texture)

bathy_png <- tempfile(fileext = ".png")

# Camera angles, shared with Panels D and E so the 3D scenes face the same way:
# view_theta is the compass azimuth, view_phi the elevation above the horizon. A
# shallow view_phi keeps the eye low and near side-on, which opens up the gaps
# between the exploded raster layers in Panel D.
view_theta <- 35
view_phi   <- 18

# zscale trades horizontal against vertical scale. After the 12x upsampling the
# matrix is 120 cells wide and the seafloor reaches ~400 m, so a smaller zscale
# than the raw 10-cell grid would use keeps the relief readable rather than a
# spike. Tune up to flatten, down to exaggerate.
rayshader::plot_3d(
  bathy_shade, elev_mat,
  zscale      = 8,
  solid       = TRUE,  solidcolor = "grey30",
  water       = TRUE,  waterdepth = 0,
  watercolor  = "#7fb2e5", wateralpha = 0.45,
  theta = view_theta, phi = view_phi, zoom = 0.72,
  windowsize  = c(900, 700)
)
rayshader::render_snapshot(filename = bathy_png, clear = TRUE)

# Read the render back as a grob for the composite figure below. wrap_elements()
# accepts a title via labs(), so the panel matches the A/C headers.
bathy_grob <- grid::rasterGrob(png::readPNG(bathy_png), interpolate = TRUE)
p_bathy <- patchwork::wrap_elements(full = bathy_grob) +
  labs(title = "B. Bathymetry") +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", size = 12))
```

### Panel C: standard depths

The standard depths are one-dimensional — a set of levels down the water
column. We draw them as points on an inverted depth axis so the surface sits at
the top, matching how the 3D scene is oriented.

```{r plot-depths}
depth_df <- data.frame(depth = depths)

p_depths <- ggplot(depth_df, aes(x = 0, y = depth)) +
  geom_point(colour = "#2c7fb8", size = 2) +
  geom_text(aes(label = paste0(depth, " m")), hjust = -0.25, size = 3) +
  scale_y_reverse(limits = c(max(depths), min(depths) - 10)) +
  scale_x_continuous(limits = c(-0.1, 1)) +
  labs(title = "C. Standard depths") +
  theme_voxel_input() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
p_depths
```

## Assembling the voxel model

[create_study_voxel()] takes the three inputs and returns a `study_voxel`. It
projects the bathymetry onto the template, flips it to a positive seafloor
depth (clamping land to 0), and stores the sorted standard depths alongside the
empty grid.

```{r create-voxel}
voxel <- create_study_voxel(
  template   = template,
  bathymetry = bathymetry,
  depths     = depths
)
voxel
#> <study_voxel>
#>   grid:   10 x 10 cells
#>   depths: 10 levels (0 to 400 m)

# The three components the object carries:
voxel$grid        # empty horizontal template
voxel$seafloor    # positive seafloor depth on that grid
voxel$depths      # sorted standard depth levels
```

## Visualizing the output: the voxelized ocean

The `study_voxel` describes a stack of cubes: at every ocean cell, the water
column from the surface (0 m) down to the seafloor is sliced into voxels at the
standard depths, with the deepest voxel clamped to the local seafloor. That is
exactly the volume [voxelize_range()] fills for a feature and that
[calc_volume()] and [calc_volume_overlap()] measure.

We render it with `rayrender`, which builds a path-traced scene from real 3D
primitives — one solid, opaque plate per voxel. First we expand the `study_voxel`
into a table of voxels (one row per ocean cell x standard-depth layer), then draw
each row as a solid plate coloured by depth. We show the result two ways: **Panel
D** pulls the standard-depth layers apart into an *exploded* stack so each raster
layer is legible on its own, and **Panel E** drops those same plates to their
true depths, so the real vertical spacing of the voxelized ocean — fine near the
surface, widening below — is visible.

```{r voxel-columns}
# Vertical extent of each standard-depth layer: half-way to the neighbouring
# levels, so adjacent voxels meet without gaps or overlap. The shallowest layer
# starts at the surface (0 m); the deepest extends half a step below the last
# level before being clamped to the seafloor per cell.
depth_bands <- function(standard_depths) {
  sd <- sort(unique(standard_depths))
  n  <- length(sd)
  mids <- (sd[-n] + sd[-1]) / 2
  data.frame(
    depth  = sd,
    z_low  = c(0, mids),
    z_high = c(mids, sd[n] + (sd[n] - sd[n - 1]) / 2)
  )
}

# Expand a study_voxel into one row per (ocean cell x depth layer) whose top
# sits above the seafloor, clamping the deepest layer to the local seafloor.
voxel_columns <- function(voxel) {
  bands <- depth_bands(voxel$depths)
  cells <- terra::as.data.frame(voxel$seafloor, xy = TRUE, na.rm = FALSE)

  do.call(rbind, lapply(seq_len(nrow(cells)), function(i) {
    sfd <- cells$seafloor[i]
    if (is.na(sfd) || sfd <= 0) return(NULL)      # skip land / dry cells
    keep <- bands$z_low < sfd
    if (!any(keep)) return(NULL)
    b <- bands[keep, , drop = FALSE]
    b$z_high <- pmin(b$z_high, sfd)               # clamp deepest voxel to seafloor
    data.frame(x = cells$x[i], y = cells$y[i],
               depth = b$depth, z_low = b$z_low, z_high = b$z_high)
  }))
}

vox <- voxel_columns(voxel)
nrow(vox)   # number of voxels (ocean cells x layers)
#> [1] 558
```

`rayrender`'s vertical axis is **y** (up), so we map the horizontal grid to the
scene's x/z plane and depth to y, going negative downward: the sea surface sits
at `y = 0` and the seafloor below it. Depth in metres dwarfs the 10-unit-wide
grid, so we scale it (`vscale`) to keep the columns readable rather than
needle-thin. Each voxel is a **solid, opaque plate** (`rayrender::cube()`)
coloured by depth; a small horizontal seam keeps the individual raster cells
legible. A single `voxel_plate()` helper places a plate at any vertical position,
so the *same* voxels can be drawn two ways: exploded onto evenly spaced levels
(Panel D) or dropped to their true standard depths (Panel E). Each column's
deepest plate is clamped to the seafloor, so the plate bottoms already trace the
seabed without a filled surface. We point the camera down Panel B's view angles,
so all three 3D panels face the same way.

```{r voxel-scene}
# Depth -> colour: shallow pale blue, deep navy.
depth_palette <- grDevices::colorRampPalette(
  c("#bcdcf5", "#3a7fc0", "#0a2f66")
)(256)
depth_to_colour <- function(d, dmin, dmax) {
  idx <- pmax(1, pmin(256, ceiling((d - dmin) / (dmax - dmin) * 256)))
  depth_palette[idx]
}

res     <- terra::res(voxel$grid)       # horizontal cell size (x, y) in grid units
dx      <- res[1]
dy      <- res[2]
vscale  <- 6 / max(vox$z_high)          # true depths -> ~6 scene units at the seafloor
hgap    <- 0.94                         # horizontal seams so raster cells stay visible
plate_h <- 0.07                         # thickness of each solid layer plate
dmin    <- min(voxel$depths)
dmax    <- max(voxel$depths)

# Layer index of each voxel (1 = surface layer), used to explode the stack.
layer_levels <- sort(unique(voxel$depths))
vox$layer <- match(vox$depth, layer_levels)

# One solid, opaque plate per voxel, coloured by depth. `ycentre` sets where the
# plate sits vertically; every plate is the same thin thickness, so the same
# voxels can be drawn exploded (evenly spaced) or at their true depths.
voxel_plate <- function(r, ycentre) {
  rayrender::cube(
    x = r$x, y = ycentre, z = r$y,
    xwidth = dx * hgap, ywidth = plate_h, zwidth = dy * hgap,
    material = rayrender::diffuse(color = depth_to_colour(r$depth, dmin, dmax))
  )
}

# Panel D (exploded): each plate is pulled to an evenly spaced level by its layer
# index, opening space between the raster layers so each is legible on its own
# and the footprint (shrinking with depth) is easy to read. Panel D gets its own
# tall column in the figure, so the spacing is exaggerated generously.
explode_step <- 2.0                     # vertical gap between exploded layers
exploded_cubes <- lapply(seq_len(nrow(vox)), function(i) {
  r <- vox[i, ]
  voxel_plate(r, -(r$layer - 1) * explode_step)
})

# Panel E (stacked): the same plates dropped to their true standard depths
# (clamped to the seafloor for each cell's deepest layer), so the real vertical
# spacing shows — fine near the surface, widening below. The plates' clamped
# bottoms already trace the seabed, so no solid staircase surface is drawn.
stacked_cubes <- lapply(seq_len(nrow(vox)), function(i) {
  r <- vox[i, ]
  voxel_plate(r, -pmin(r$depth, r$z_high) * vscale)
})

# A soft overhead light, high and toward the camera, shared by both scenes.
key_light <- rayrender::sphere(
  x = 14, y = 20, z = -10, radius = 3,
  material = rayrender::light(intensity = 8)
)

# Assemble the two scenes: both are just the plates and a shared light, floating
# on the plain white background.
scene_exploded <- Reduce(rayrender::add_object, exploded_cubes) |>
  rayrender::add_object(key_light)

scene_stacked <- Reduce(rayrender::add_object, stacked_cubes) |>
  rayrender::add_object(key_light)
```

```{r voxel-render}
library(png)

# Camera matched to Panel B. rayshader hands theta/phi to rgl, whose eye lies
# along (cos(phi)sin(theta), sin(phi), cos(phi)cos(theta)) in (east, up, south) —
# rgl's default view looks north, so its +z is the south side. Our scene uses
# +z = north, so we negate that term. This unit direction puts the deep
# south-east corner in front, exactly as in Panel B.
rad     <- pi / 180
cam_dir <- c(
   cos(view_phi * rad) * sin(view_theta * rad),   # east
   sin(view_phi * rad),                           # up
  -cos(view_phi * rad) * cos(view_theta * rad)    # north (rgl south -> negate)
)

# rayshader and rayrender lay out their horizontal axes with opposite handedness,
# so a raw rayrender frame comes out mirrored east-west relative to Panel B.
# Reflecting the finished image left-right (equivalent to reflecting camera and
# scene together, which keeps the deep corner in front) restores the match.
flip_h <- function(img) img[, rev(seq_len(dim(img)[2])), , drop = FALSE]

# Both output panels share one camera (same distance, fov and tall portrait
# frame), so D and E render at exactly the same zoom -- the 10x10 grid footprint
# reads at the same size in both. Only the aim (`lookat_y`) differs, centring
# each stack: D's exploded layers fill the frame, while E's shorter true-depth
# stack sits lower and reads as smaller at the same scale.
render_voxels <- function(scene, file, lookat_y, dist, fov, width, height) {
  rayrender::render_scene(
    scene,
    lookfrom       = c(5, 0, 5) + dist * cam_dir,
    lookat         = c(5, lookat_y, 5),
    fov            = fov,
    samples        = 256,
    width          = width, height = height,
    ambient_light  = TRUE,
    backgroundhigh = "#ffffff", backgroundlow = "#ffffff",  # solid white, matches the figure
    filename       = file
  )
  flip_h(png::readPNG(file))
}

exploded_png <- tempfile(fileext = ".png")
stacked_png  <- tempfile(fileext = ".png")

# Panel D: exploded plates span 0 to -(nlayers - 1) * explode_step; aim at its mid.
exploded_mid <- -(length(layer_levels) - 1) * explode_step / 2
img_exploded <- render_voxels(scene_exploded, exploded_png,
                              lookat_y = exploded_mid, dist = 42, fov = 30,
                              width = 880, height = 1200)
# Panel E: same dist/fov/frame as D (so the zoom matches); aimed at its own
# shorter span (0 to ~-6, mid ~-3), which leaves it sitting lower in the frame.
img_stacked  <- render_voxels(scene_stacked, stacked_png,
                              lookat_y = -3, dist = 42, fov = 30,
                              width = 880, height = 1200)

p_output1 <- patchwork::wrap_elements(
  full = grid::rasterGrob(img_exploded, interpolate = TRUE)) +
  labs(title = "D. Exploded raster stack") +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", size = 12))

p_output2 <- patchwork::wrap_elements(
  full = grid::rasterGrob(img_stacked, interpolate = TRUE)) +
  labs(title = "E. Voxelized ocean") +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", size = 12))
```

Panel D pulls the raster layers apart so each standard-depth slice reads on its
own, its footprint shrinking with depth as the rising seafloor drops out cells;
Panel E drops those same plates to their true depths — spaced finely near the
surface and widening below — their clamped bottoms tracing the seabed that Panel
B renders as a surface: short over the shallow shelf and reaching deep in the
south-east, absent over the north-west land. These plates are the same voxels [voxelize_range()] fills for a
feature and that [calc_volume()] and [calc_volume_overlap()] measure. Because
`rayrender` path-traces real geometry, the render is deterministic and needs no
browser or `webshot` step — `render_scene()` writes the PNG directly.

## Composing the manuscript figure

The manuscript figure is three columns wide: the three inputs (A, B, C) stacked
on the left, the exploded raster stack (D) in the middle, and the voxelized ocean
(E) on the right. All four 3D panels are PNG renders wrapped as grobs (`p_bathy`,
`p_output1`, `p_output2`); the two 2D panels (`p_template`, `p_depths`) are
ggplots. The inputs column is kept narrow (its plots are small), while D and E
get equal widths and share one camera, so they read at the same size and zoom.

```{r compose-figure, fig.width = 14, fig.height = 8}
inputs_col <- (p_template / p_bathy / p_depths)

fig_voxel_model <- inputs_col | p_output1 | p_output2
fig_voxel_model <- fig_voxel_model +
  plot_layout(widths = c(0.7, 1, 1)) 
fig_voxel_model

# Save into this figures/ directory (this document lives here, and render runs
# with figures/ as the working directory).
ggsave(
  filename = "fig-voxel-model_raw.pdf",
  plot = fig_voxel_model,
  width = 14, height = 8, dpi = 300, bg = "white"
)

# Figures modified manually in vector editing software (Inkscape) after saving.
```

`fig-voxel-model_raw.pdf` is the raw export; it is hand-edited in Inkscape to
produce the final `fig-voxel-model.pdf` / `fig-voxel-model.png` used in the
manuscript. All three live here in `figures/` alongside this source.

## Recap

- The voxel model is a stack of raster layers sharing one grid, each layer a
  standard depth — the same representation oceanographic products use.
- [create_study_voxel()] bundles the three inputs every 3D operation needs:
  the **horizontal grid template**, the **bathymetry** that caps each column,
  and the **standard depths** that slice the water column.
- The output is a voxelized ocean: per-cell columns of cubes clamped to the
  seafloor, which [voxelize_range()], [calc_volume()], and
  [calc_volume_overlap()] operate on directly.
