---
title: "3D Volume Overlap Between Species and Fisheries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3D Volume Overlap Between Species and Fisheries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 3D Volume Overlap Between Species and Fisheries

This vignette recreates the Bangladesh fisheries analysis (Haque et al.),
which quantified the three-dimensional overlap between critically endangered
shark and ray species and artisanal fisheries in Bangladesh waters. The key
insight from the original study was that horizontal (2D) overlap between
species ranges and fisheries can be large, but the actual 3D volume overlap
is often much smaller because species occupy depth ranges that provide
refuge from fishing gear.

The original analysis used a hexagonal grid, custom depth-overlap
calculations, and participatory mapping data. Here we reproduce the same
methodology using the `sharkabc3d` raster-based volume functions.

## Overview

The workflow has four steps:

1. **Load inputs**: species ranges, fishery footprints, bathymetry
2. **Rasterize** each species and fishery onto the bathymetry grid
3. **Calculate 3D volume overlap** between each species-fishery pair
4. **Visualise** cumulative pressure and overlap by depth

## Setup

```{r setup}
# devtools::load_all()
library(sharkabc3d)
library(dplyr)
library(stringr)
library(readxl)
library(here)

sf::sf_use_s2(FALSE)
```

## Step 1: Load Input Data

### 1a: Bathymetry and EEZ

The original analysis used GEBCO bathymetry for the Bangladesh EEZ at 1 km
resolution. `load_gebco_bathymetry()` loads and standardises the raster (positive
depth values, in metres).

```{r load-layers}
# Load bathymetry
bathy <- load_gebco_bathymetry(
  "~/Programming_Projects/Big_Data/gebco_2025_sub_ice_topo/GEBCO_2025_sub_ice.nc"
)

# Load Bangladesh EEZ
bangladesh_eez <- sf::st_read(
  "~/Programming_Projects/Big_Data/World_EEZ_v12_20231025/eez_v12.shp",
  query = "SELECT * FROM eez_v12 WHERE TERRITORY1 = 'Bangladesh'",
  quiet = TRUE
) %>% rename(geometry = "_ogr_geometry_")
```

### 1b: Species ranges with depth limits

We read the study species list from the Excel file, query the IUCN Red List
`SHARKS_RAYS_CHIMAERAS` shapefile for their global ranges, crop to the
Bangladesh EEZ, then merge with the depth limits provided in the study and
the IUCN depth limits fetched via the API.

```{r load-species}
# Load species depth information from Excel
excel_path <- here(
  "example_data",
  "Maps for Jay_3D overlap", "Spp&Fisherydata_Jay.xlsx"
)

species_info <- read_excel(excel_path, sheet = "Spp. depth range") %>%
  rename(
    scientific_name = `Valid scientific name`,
    depth_min_haque = Lowest_depth_m,
    depth_max_haque = Highest_depth_m
  )

# Query the IUCN Red List shapefile for the study species.
# Filter: presence = 1 (extant), origin in (1, 2) (native/reintroduced),
# seasonal in (1, 2, 3) (resident/breeding/non-breeding).
iucn_shp <- "~/Programming_Projects/Big_Data/SHARKS_RAYS_CHIMAERAS/SHARKS_RAYS_CHIMAERAS.shp"
layer_name <- sf::st_layers(iucn_shp)$name[1]

sp_list_sql <- paste0("'", species_info$scientific_name, "'", collapse = ", ")
sp_query <- str_glue(
  "SELECT sci_name, presence, origin, seasonal FROM \"{layer_name}\"",
  " WHERE sci_name IN ({sp_list_sql})",
  " AND presence IN (1) AND origin IN (1, 2) AND seasonal IN (1, 2, 3)"
)

species_global <- sf::st_read(iucn_shp, query = sp_query, quiet = TRUE) %>%
  sf::st_transform(4326)

# Crop to Bangladesh EEZ and dissolve multipolygons to one geometry per species
species_ranges <- species_global %>%
  sf::st_make_valid() %>%
  sf::st_intersection(bangladesh_eez %>% select(geometry)) %>%
  group_by(sci_name) %>%
  summarise(geometry = sf::st_union(geometry), .groups = "drop")

message("Loaded ", nrow(species_ranges), " species ranges from IUCN")

# Fetch IUCN depth limits via API for species that have IUCN data
api_key <- Sys.getenv("IUCN_REDLIST_KEY")
species_iucn <- fetch_species_assessments(
  api_key, species_names = species_ranges$sci_name
)
species_iucn <- species_iucn %>%
  mutate(fill_missing_depths(upper_depth_limit, lower_depth_limit, genus_name))

# Merge ranges + IUCN depths + provided depths into one sf object
bangladesh_species <- species_ranges %>%
  left_join(
    species_iucn %>% select(
      scientific_name,
      depth_min_iucn = upper_depth,
      depth_max_iucn = lower_depth
    ),
    by = c("sci_name" = "scientific_name")
  ) %>%
  left_join(
    species_info,
    by = c("sci_name" = "scientific_name")
  ) %>%
  mutate(
    depth_min_sel = dplyr::coalesce(depth_min_haque, depth_min_iucn),
    depth_max_sel = dplyr::coalesce(depth_max_haque, depth_max_iucn)
  )
```

### 1c: Fishery footprints with depth ranges

The original study used participatory mapping (fisher interviews + KDE
heatmaps) to define the spatial footprints and operating depth ranges of
7 artisanal sub-fisheries. These are loaded as polygons with associated
gear depth ranges.

```{r load-fisheries}
# Load fishery depth information from Excel
# Tidal correction: GEBCO bathymetry uses mean sea level (MSL) as its vertical
# datum, but Bangladesh coastal waters experience 4-6 m spring tidal range.
# Fishers operate on favourable tides, so a cell at e.g. 4 m MSL depth may
# have 7 m of water at high tide — within a fishery's operating range.
# Subtracting half the mean tidal range (~3 m) from the fishery depth_min
# accounts for the tidal envelope that static MSL bathymetry misses.
tidal_correction_m <- 3

fishery_info <- read_excel(excel_path, sheet = "Fishery_depth range") %>%
  rename(
    fishery_type = `Fishery type`,
    depth_min = `Depthrange_low_(m)`,
    depth_max = `Depthrange_high_(m)`
  ) %>%
  mutate(depth_min = pmax(depth_min - tidal_correction_m, 0))

# Load fishery footprint shapefiles
fishery_shp_dir <- here(
  "example_data",
  "Maps for Jay_3D overlap", "Fisheries foot print_shapefiles"
)

# Map fishery type codes to shapefile subdirectory names
fishery_shp_names <- c(
  SGN = "SGN_eez_intersect",
  MGN = "MGN",
  LGN = "LGN_eez",
  ULL = "ULL_eez",
  BLL = "BLL_eez",
  SBN = "SBN_eez",
  PTN = "PTN_eez"
)

fishery_footprints <- lapply(fishery_info$fishery_type, function(ft) {
  shp_name <- fishery_shp_names[[ft]]
  sf::st_read(
    file.path(fishery_shp_dir, shp_name, paste0(shp_name, ".shp")),
    quiet = TRUE
  )
})
names(fishery_footprints) <- fishery_info$fishery_type

# Combine footprints into single sf, crop to the Bangladesh EEZ, and join
# depth info. Project all to WGS84 (EPSG:4326) before binding since the raw
# shapefiles have different CRS. The EEZ intersection ensures fishery extents
# never extend past the analysis boundary.
fisheries <- lapply(names(fishery_footprints), function(ft) {
  fishery_footprints[[ft]] %>%
    sf::st_transform(4326) %>%
    sf::st_make_valid() %>%
    sf::st_union() %>%
    sf::st_as_sf() %>%
    mutate(fishery_type = ft)
}) %>%
  bind_rows() %>%
  sf::st_intersection(bangladesh_eez %>% select(geometry)) %>%
  left_join(fishery_info, by = "fishery_type")
```

## Step 2: Create Study Area Grid and Rasterize Layers

### 2a: Create study area raster

`create_study_raster()` computes a combined extent from a list of spatial
objects and returns an empty raster grid at the desired resolution and CRS.

```{r study-raster}
# Lambert Azimuthal Equal Area centered on Bangladesh waters
bangladesh_crs <- "+proj=laea +lat_0=22 +lon_0=90 +datum=WGS84 +units=m"

study_grid <- create_study_raster(
  layers = list(bangladesh_eez),
  res = 1000,
  crs = bangladesh_crs
)

# Crop bathymetry to study area before projecting (avoids re-projecting
# the full GEBCO raster). Use the WGS84 EEZ polygon directly for cropping
# since reprojecting the LAEA grid to WGS84 gives a near-global extent.
bathy_crop <- terra::crop(bathy, terra::vect(bangladesh_eez))
seafloor <- terra::project(bathy_crop, study_grid)
seafloor <- terra::clamp(-seafloor, lower = 0)
```

### 2b: Rasterize species

`voxelize_range()` converts each polygon + depth range into a multi-layer
SpatRaster (presence, depth_min, depth_max) aligned to the study grid.
The depth_max is clamped to the seafloor per cell so that no species or
fishery extends deeper than the actual bathymetry.

```{r rasterize-species}
species_rasters <- voxelize_ranges(
  sf_data = bangladesh_species,
  voxel = study_grid,
  bathymetry = seafloor,
  depth_min_col = "depth_min_sel",
  depth_max_col = "depth_max_sel",
  name_col = "sci_name"
)
```

### 2c: Rasterize fisheries

```{r rasterize-fisheries}
fishery_rasters <- voxelize_ranges(
  sf_data = fisheries,
  voxel = study_grid,
  bathymetry = seafloor,
  depth_min_col = "depth_min",
  depth_max_col = "depth_max",
  name_col = "fishery_type"
)
```

## Step 3: Calculate 3D Volume Overlap

For each species-fishery pair, `calc_volume_overlap()` computes the
per-cell depth overlap and returns both the overlap raster and a summary
with the overlap volume and proportion of each range's total volume.

### 3a: Single species-fishery example

```{r single-overlap}
# Example: overlap between first species and first fishery
overlap_rast <- calc_volume_overlap(
  range_rast_a = species_rasters[[1]],
  range_rast_b = fishery_rasters[[1]]
)

message(
  "Overlap between ", names(species_rasters)[1],
  " and ", names(fishery_rasters)[1]
)

# Per-cell volume raster stack (volume_a, volume_b, volume_overlap in km³)
overlap_rast

# Total volumes
terra::global(overlap_rast, "sum", na.rm = TRUE)
```

### 3b: All species-fishery combinations

```{r all-overlaps}
# Compute overlap for every species x fishery pair
pairs <- expand.grid(
  species = names(species_rasters),
  fishery = names(fishery_rasters),
  stringsAsFactors = FALSE
)

n_pairs <- nrow(pairs)
message("Computing ", n_pairs, " species x fishery overlaps...")
pb <- txtProgressBar(min = 0, max = n_pairs, style = 3)

# Store rasters and summary table
overlap_rasters <- vector("list", n_pairs)
overlap_summaries <- vector("list", n_pairs)

for (i in seq_len(n_pairs)) {
  sp <- pairs$species[i]
  fi <- pairs$fishery[i]

  ov <- calc_volume_overlap(
    range_rast_a = species_rasters[[sp]],
    range_rast_b = fishery_rasters[[fi]]
  )

  overlap_rasters[[i]] <- ov
  totals <- terra::global(ov, "sum", na.rm = TRUE)
  overlap_summaries[[i]] <- tibble::tibble(
    species = sp,
    fishery = fi,
    volume_species_km3 = totals["volume_a", "sum"],
    volume_fishery_km3 = totals["volume_b", "sum"],
    volume_overlap_km3 = totals["volume_overlap", "sum"]
  )

  setTxtProgressBar(pb, i)
}
close(pb)

names(overlap_rasters) <- paste(pairs$species, pairs$fishery, sep = " x ")
overlap_results <- bind_rows(overlap_summaries)
head(overlap_results)
```

### 3c: Depth refuge — area overlap vs volume overlap

The key finding of the original study: horizontal (2D) overlap between a
species range and a fishery can be large, but the actual 3D volume overlap
is much smaller when the species occupies depths below the fishing gear.
Here we quantify that **depth refuge** for every species × fishery pair by
comparing the horizontal area overlap with the volume overlap.

For each pair we compute the proportion of the species exposed to the
fishery two ways: `area_overlap_prop` (share of the species' horizontal
footprint that the fishery covers) and `volume_overlap_prop` (share of the
species' 3D volume the fishery actually overlaps). The gap between them is
the depth refuge: `depth_refuge = 1 - volume_overlap_prop / area_overlap_prop`
(0 = no refuge, the 3D overlap matches the 2D footprint; → 1 = strong
refuge, the species shares the map with the fishery but sits below it).

```{r depth-refuge}
# For each species x fishery pair, compare the 2D (area) overlap with the
# 3D (volume) overlap to quantify depth refuge.
depth_refuge <- lapply(seq_len(n_pairs), function(i) {
  sp <- pairs$species[i]
  fi <- pairs$fishery[i]
  ov <- overlap_rasters[[i]]

  cell_area_km2 <- terra::cellSize(ov[["depth_min_a"]], unit = "km")

  # Horizontal footprints (presence, ignoring depth)
  sp_present   <- !is.na(ov[["depth_min_a"]])
  both_present <- sp_present & !is.na(ov[["depth_min_b"]])

  # 2D areas (km^2): species footprint and species x fishery intersection
  species_area_km2 <- terra::global(
    terra::mask(cell_area_km2, terra::ifel(sp_present, 1, NA)),
    "sum", na.rm = TRUE)$sum
  overlap_area_km2 <- terra::global(
    terra::mask(cell_area_km2, terra::ifel(both_present, 1, NA)),
    "sum", na.rm = TRUE)$sum

  # 3D volumes (km^3): species volume and species x fishery overlap volume
  totals <- terra::global(ov, "sum", na.rm = TRUE)
  species_volume_km3 <- totals["volume_a", "sum"]
  overlap_volume_km3 <- totals["volume_overlap", "sum"]

  # Exposure of the species to the fishery, measured two ways
  area_overlap_prop   <- overlap_area_km2 / species_area_km2
  volume_overlap_prop <- overlap_volume_km3 / species_volume_km3

  tibble::tibble(
    species = sp,
    fishery = fi,
    species_area_km2 = species_area_km2,
    overlap_area_km2 = overlap_area_km2,
    species_volume_km3 = species_volume_km3,
    overlap_volume_km3 = overlap_volume_km3,
    area_overlap_prop = area_overlap_prop,
    volume_overlap_prop = volume_overlap_prop,
    # Depth refuge: fraction of the horizontal exposure removed once depth is
    # accounted for. 0 = no refuge; -> 1 = strong depth refuge.
    depth_refuge = ifelse(
      area_overlap_prop > 0,
      1 - volume_overlap_prop / area_overlap_prop,
      NA_real_
    )
  )
}) %>% bind_rows()

# Save the per-pair area-vs-volume comparison to a CSV under tmp/
tmp_dir <- here("tmp")
dir.create(tmp_dir, showWarnings = FALSE, recursive = TRUE)
depth_refuge_csv <- file.path(tmp_dir, "depth_refuge_area_vs_volume.csv")
write.csv(depth_refuge, depth_refuge_csv, row.names = FALSE)
message("Wrote depth refuge comparison to ", depth_refuge_csv)

# Pairs with the greatest depth refuge: large horizontal overlap but small
# volume overlap (species sits below the fishing gear).
depth_refuge %>%
  filter(overlap_area_km2 > 0) %>%
  arrange(desc(depth_refuge)) %>%
  select(species, fishery, area_overlap_prop, volume_overlap_prop, depth_refuge) %>%
  head(10)
```

## Step 4: Comparison with Original Analysis

Having computed the 3D overlaps, we now compare our raster-based results
against the reference outputs from the original analysis (Haque et al.),
which used a hexagonal grid approach. We expect `sharkabc3d` volumes to be
systematically lower for two reasons:

1. **Per-cell bathymetry clamping**: The original analysis applied a single
   uniform depth range across the entire intersection (volume = area ×
   depth_range). `sharkabc3d` clamps `depth_max` to the actual seafloor in
   each cell, so cells shallower than the nominal depth range contribute less
   volume. On the shallow Bangladesh continental shelf, this significantly
   reduces effective depth.
2. **Raster vs hexagonal grid**: Minor differences in spatial footprint from
   rasterization onto a square grid vs the original hexagonal grid.

> **Note:** This step compares against reference results from the original
> Haque et al. analysis. That CSV is not distributed with the package —
> contact the maintainer for it, and place it under `example_data/` alongside
> the other inputs used above. The analysis (Steps 1–3) and the visualisation
> below (Step 5) run without it.

```{r compare-original}
# Load reference results from original analysis (per species-fishery pair)
reference <- read.csv(
  here("example_data", "bangladesh-fisheries-3d-overlap",
       "230314_3D_species_fisheries_intersect.csv")
) %>%
  rename(species = genus_species)

# Join our results to the reference
comparison <- overlap_results %>%
  left_join(reference, by = c("species", "fishery"))

# Compute our area and mean depth for the overlap to compare with reference
# (calc_volume_overlap returns volume rasters; extract area and depth separately)
overlap_diagnostics <- lapply(seq_len(nrow(pairs)), function(i) {
  sp <- pairs$species[i]
  fi <- pairs$fishery[i]
  ov <- overlap_rasters[[i]]

  # Overlap cells: where volume_overlap > 0
  ov_vol <- ov[["volume_overlap"]]
  ov_present <- !is.na(ov_vol) & ov_vol > 0

  # Count overlapping cells and their area
  cell_area_m2 <- terra::cellSize(ov_vol, unit = "m")
  area_rast <- terra::mask(cell_area_m2, terra::ifel(ov_present, 1, NA))
  our_area_m2 <- terra::global(area_rast, "sum", na.rm = TRUE)$sum

  # Total overlap volume in m³ for direct comparison
  our_volume_m3 <- terra::global(ov_vol, "sum", na.rm = TRUE)$sum * 1e9

  tibble::tibble(
    species = sp,
    fishery = fi,
    our_intersect_area_m2 = our_area_m2,
    our_intersect_volume_m3 = our_volume_m3
  )
}) %>% bind_rows()

comparison <- comparison %>%
  left_join(overlap_diagnostics, by = c("species", "fishery")) %>%
  mutate(
    area_ratio = our_intersect_area_m2 / intersect_area_m2,
    volume_ratio = our_intersect_volume_m3 / intersect_volume_m3,
    volume_diff_pct = (our_intersect_volume_m3 - intersect_volume_m3) /
      intersect_volume_m3 * 100
  )

# Summary of agreement
message("Matched pairs: ", sum(!is.na(comparison$intersect_volume_m3)),
        " / ", nrow(comparison))

message("\nArea ratio (ours / original):")
summary(comparison$area_ratio)

message("\nVolume ratio (ours / original):")
summary(comparison$volume_ratio)

# Is the discrepancy driven by area, depth, or both?
comparison <- comparison %>%
  mutate(
    # Implied mean depth of overlap: volume / area
    ref_mean_depth_m = intersect_volume_m3 / intersect_area_m2,
    our_mean_depth_m = our_intersect_volume_m3 / our_intersect_area_m2,
    depth_ratio = our_mean_depth_m / ref_mean_depth_m
  )

message("\nMean overlap depth ratio (ours / original):")
summary(comparison$depth_ratio)
```

```{r compare-plots, fig.width=8, fig.height=4}
par(mfrow = c(1, 3))

# Area comparison
plot(comparison$intersect_area_m2 / 1e6, comparison$our_intersect_area_m2 / 1e6,
     xlab = "Original area (km²)", ylab = "sharkabc3d area (km²)",
     main = "Intersection Area",
     pch = 19, col = adjustcolor("steelblue", 0.6))
abline(0, 1, lty = 2, col = "red")

# Mean depth comparison
plot(comparison$ref_mean_depth_m, comparison$our_mean_depth_m,
     xlab = "Original mean depth (m)", ylab = "sharkabc3d mean depth (m)",
     main = "Mean Overlap Depth",
     pch = 19, col = adjustcolor("steelblue", 0.6))
abline(0, 1, lty = 2, col = "red")

# Volume comparison
plot(comparison$intersect_volume_m3 / 1e9, comparison$our_intersect_volume_m3 / 1e9,
     xlab = "Original volume (km³)", ylab = "sharkabc3d volume (km³)",
     main = "Overlap Volume",
     pch = 19, col = adjustcolor("steelblue", 0.6))
abline(0, 1, lty = 2, col = "red")

par(mfrow = c(1, 1))
```

| Aspect | Original (Haque et al.) | `sharkabc3d` |
|--------|-------------------------|--------------|
| Grid type | Hexagonal grid (1 km cells) | Raster grid (1 km cells) |
| Depth handling | Uniform depth range across entire intersection footprint | Per-cell depth clamped to bathymetry via `voxelize_range()` |
| Volume formula | area × depth_range (single depth per pair) | Sum of per-cell (cell_area × clamped depth) via `calc_volume_overlap()` |
| Fishery/species richness maps | Custom ggplot code | `count_3d_overlaps()` + ggplot2 |
| Species × fishery overlap maps | Custom ggplot code | `calc_volume_overlap()` + ggplot2/patchwork |
| **Reproducibility** | **Single-use scripts** | **Reusable for any species x fishery** |

`sharkabc3d` produces lower volume estimates than the original analysis
(median volume ratio ~0.33). The difference is driven primarily by per-cell
bathymetry clamping (~56% of the reduction) and secondarily by raster vs
hexagonal grid edge effects (~25%). The per-cell approach is more physically
accurate: cells on the shallow continental shelf where the seafloor is
shallower than the nominal depth range contribute proportionally less volume,
rather than being assigned the full depth range of the species-fishery pair.

## Step 5: Visualisation

This step recreates the three sets of maps from the original analysis:

- **Fishery richness per species**: For each species, how many fisheries
  have 3D overlap at each grid cell
- **Species richness per fishery**: For each fishery, how many species
  have 3D overlap at each grid cell
- **Individual species × fishery overlap**: Map of each range and their
  intersection, with a depth profile sidebar

### 5a: Prepare map context and helpers

```{r vis-setup}
library(ggplot2)
library(tidyterra)
library(ggnewscale)
library(patchwork)

# Use the Bangladesh EEZ polygon itself for map context — it has the exact
# same geometry used for the species-range intersection, so the boundary
# drawn on the map lines up with the analysis extent.
plot_bbox <- sf::st_bbox(bangladesh_eez)

# Build a land mask from the plot bounding box by subtracting EEZ ocean areas.
# Load neighbouring EEZs (India, Myanmar) so the full ocean extent is removed.
neighbour_eez <- sf::st_read(

  "~/Programming_Projects/Big_Data/World_EEZ_v12_20231025/eez_v12.shp",
  query = "SELECT * FROM eez_v12 WHERE TERRITORY1 IN ('India', 'Myanmar')",
  quiet = TRUE
) %>% rename(geometry = "_ogr_geometry_")

all_eez <- sf::st_union(
  sf::st_geometry(bangladesh_eez),
  sf::st_union(sf::st_geometry(neighbour_eez))
)

bbox_poly <- sf::st_as_sfc(plot_bbox)
land_mask <- sf::st_difference(bbox_poly, all_eez)

# Remove the furthest-south polygon (ocean gap between EEZ boundaries)
land_parts <- sf::st_cast(land_mask, "POLYGON") %>% sf::st_as_sf()
centroids_y <- sf::st_coordinates(sf::st_centroid(land_parts))[, "Y"]
land_mask <- sf::st_union(land_parts[-which.min(centroids_y), ]) 

# Colour palette for species/fishery/intersect maps and depth profiles
col_species   <- "#440154"
col_intersect <- "#21918c"
col_fishery   <- "#fde725"

theme_map <- function() {
  theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
    )
}
```

### 5b: Fishery richness per species

For each species, a map showing how many fisheries have 3D volume overlap
at each grid cell. Cells are coloured only where the species is present
and at least one fishery overlaps in depth. Recreates the
`all_fishery_by_species/` plots from the original analysis.

```{r fishery-richness-maps, fig.width=18, fig.height=18}
cr_species <- species_iucn %>%
  filter(red_list_category == "CR") %>%
  pull(scientific_name)
cr_species <- intersect(cr_species, names(species_rasters))

fishery_richness_plots <- lapply(cr_species, function(sp) {
  sp_rast <- species_rasters[[sp]]

  # Stack of 1/NA overlap rasters, summed across layers, gives the number of
  # fisheries that overlap the species in 3D at each cell
  overlap_stack <- terra::rast(lapply(fishery_rasters, function(fi_rast) {
    count_3d_overlap(sp_rast, fi_rast)
  }))
  count_rast <- sum(overlap_stack, na.rm = TRUE) %>% project("EPSG:4326")

  # Species presence raster for background
  sp_presence <- terra::ifel(!is.na(sp_rast[["depth_min"]]), 1, NA) %>%
    terra::project("EPSG:4326", method = "near")

  ggplot() +
    geom_sf(data = bangladesh_eez, aes(colour = "Bangladesh EEZ"),
            fill = "white", linewidth = 0.4, linetype = "dashed") +
    geom_sf(data = land_mask, fill = "grey90", colour = "grey50", linewidth = 0.5) +
    geom_spatraster(data = sp_presence) +
    scale_fill_gradient(
      low = "#a3fff9", high = "#a3fff9", na.value = "transparent",
      guide = guide_legend(title = NULL,
                           override.aes = list(fill = "#a3fff9")),
      labels = "Species range"
    ) +
    new_scale_fill() +
    geom_spatraster(data = count_rast) +
    scale_fill_viridis_c(
      name = "# Fisheries",
      limits = c(1, length(fishery_rasters)),
      breaks = c(1, length(fishery_rasters)),
      na.value = "transparent",
      guide = guide_colorbar(title.position = "right", ticks = FALSE,
                             ticks.colour = "transparent")
    ) +
    scale_colour_manual(
      values = c("Bangladesh EEZ" = "red"),
      name = NULL
    ) +
    guides(colour = guide_legend(
      override.aes = list(fill = "white", linetype = "dashed", linewidth = 0.4)
    )) +
    coord_sf(xlim = plot_bbox[c("xmin", "xmax")],
             ylim = plot_bbox[c("ymin", "ymax")], expand = FALSE) +
    annotate("rect",
             xmin = plot_bbox[["xmin"]] + 0.02 * diff(plot_bbox[c("xmin", "xmax")]),
             xmax = plot_bbox[["xmin"]] + 0.55 * diff(plot_bbox[c("xmin", "xmax")]),
             ymin = plot_bbox[["ymax"]] - 0.05 * diff(plot_bbox[c("ymin", "ymax")]),
             ymax = plot_bbox[["ymax"]] - 0.00 * diff(plot_bbox[c("ymin", "ymax")]),
             fill = "grey90", colour = NA) +
    annotate("text",
             x = plot_bbox[["xmin"]] + 0.03 * diff(plot_bbox[c("xmin", "xmax")]),
             y = plot_bbox[["ymax"]] - 0.02 * diff(plot_bbox[c("ymin", "ymax")]),
             label = sp, fontface = "italic", hjust = 0, vjust = 1, size = 4) +
    theme_map()
})
names(fishery_richness_plots) <- cr_species

# Tile 9 random species into a 3x3 grid with shared legend
set.seed(42)
tile_species <- sample(cr_species, min(9, length(cr_species)))
tile_plots <- fishery_richness_plots[tile_species]
fishery_richness_grid <- wrap_plots(tile_plots, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 13),
        legend.key.size = unit(1, "cm"))

out_dir_fb <- here("output", "bangladesh-fisheries-3d-overlap", "fishery_by_species")
dir.create(out_dir_fb, recursive = TRUE, showWarnings = FALSE)
ggsave(file.path(out_dir_fb, "fishery_richness_grid.png"),
       fishery_richness_grid, width = 10, height = 17, dpi = 300)
lapply(names(fishery_richness_plots), function(sp) {
  ggsave(file.path(out_dir_fb, paste0(sp, ".png")),
         fishery_richness_plots[[sp]], width = 8, height = 8, dpi = 300)
})
```

### 5c: Species richness per fishery

For each fishery, a map showing how many species have 3D volume overlap
at each grid cell. Recreates the `all_species_by_fishery/` plots from the
original analysis.

```{r species-richness-maps, fig.width=18, fig.height=18}
species_richness_plots <- lapply(names(fishery_rasters), function(fi) {
  fi_rast <- fishery_rasters[[fi]]

  cr_rasters <- species_rasters[cr_species]
  overlap_stack <- terra::rast(lapply(cr_rasters, function(sp_rast) {
    count_3d_overlap(fi_rast, sp_rast)
  }))
  count_rast <- sum(overlap_stack, na.rm = TRUE) %>% project("EPSG:4326")

  ggplot() +
    geom_sf(data = bangladesh_eez, aes(colour = "Bangladesh EEZ"),
            fill = "white", linewidth = 0.4, linetype = "dashed") +
    geom_sf(data = land_mask, fill = "grey90", colour = "grey50", linewidth = 0.5) +
    geom_spatraster(data = count_rast) +
    scale_fill_viridis_c(
      name = "# Species",
      limits = c(1, length(cr_species)),
      breaks = c(1, length(cr_species)),
      na.value = "transparent",
      guide = guide_colorbar(title.position = "right", ticks = FALSE,
                             ticks.colour = "transparent")
    ) +
    scale_colour_manual(
      values = c("Bangladesh EEZ" = "red"),
      name = NULL
    ) +
    guides(colour = guide_legend(
      override.aes = list(fill = "white", linetype = "dashed", linewidth = 0.4)
    )) +
    coord_sf(xlim = plot_bbox[c("xmin", "xmax")],
             ylim = plot_bbox[c("ymin", "ymax")], expand = FALSE) +
    annotate("rect",
             xmin = plot_bbox[["xmin"]] + 0.02 * diff(plot_bbox[c("xmin", "xmax")]),
             xmax = plot_bbox[["xmin"]] + 0.55 * diff(plot_bbox[c("xmin", "xmax")]),
             ymin = plot_bbox[["ymax"]] - 0.05 * diff(plot_bbox[c("ymin", "ymax")]),
             ymax = plot_bbox[["ymax"]] - 0.00 * diff(plot_bbox[c("ymin", "ymax")]),
             fill = "grey90", colour = NA) +
    annotate("text",
             x = plot_bbox[["xmin"]] + 0.03 * diff(plot_bbox[c("xmin", "xmax")]),
             y = plot_bbox[["ymax"]] - 0.02 * diff(plot_bbox[c("ymin", "ymax")]),
             label = fi, hjust = 0, vjust = 1, size = 4) +
    theme_map()
})
names(species_richness_plots) <- names(fishery_rasters)

# Tile fisheries into a grid with shared legend (exclude LGN)
grid_plots <- species_richness_plots[setdiff(names(species_richness_plots), "LGN")]
species_richness_grid <- wrap_plots(grid_plots, ncol = 2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 13),
        legend.key.size = unit(1, "cm"))

out_dir_sb <- here("output", "bangladesh-fisheries-3d-overlap", "species_by_fishery")
dir.create(out_dir_sb, recursive = TRUE, showWarnings = FALSE)
ggsave(file.path(out_dir_sb, "species_richness_grid.png"),
       species_richness_grid, width = 10, height = 17, dpi = 300)
lapply(names(species_richness_plots), function(fi) {
  ggsave(file.path(out_dir_sb, paste0(fi, ".png")),
         species_richness_plots[[fi]], width = 8, height = 8, dpi = 300)
})
```

### 5d: Individual species × fishery overlap maps

For each species–fishery pair, a compound figure with a map panel (right)
showing species range (purple), fishery footprint (yellow), and their 3D
intersection (teal), plus a depth profile sidebar (left) showing the
number of cells at each depth occupied by the species only, fishery only,
or both. Recreates the `species_fishery_overlap_plots_png/` set.

```{r overlap-raster-filter}
# Filter to CR species only, exclude LGN fishery
cr_pairs <- pairs %>% filter(species %in% cr_species, fishery != "LGN")
cr_overlap_rasters <- overlap_rasters[pairs$species %in% cr_species & pairs$fishery != "LGN"]
```

```{r overlap-map-plots}
# Map panel for each species x fishery pair
overlap_map_plots <- lapply(seq_len(nrow(cr_pairs)), function(i) {
  sp <- cr_pairs$species[i]
  fi <- cr_pairs$fishery[i]
  ov <- cr_overlap_rasters[[i]]

  # Classify cells: 1 = species only, 2 = intersect, 3 = fishery only
  has_a <- !is.na(ov[["volume_a"]]) & ov[["volume_a"]] > 0
  has_b <- !is.na(ov[["volume_b"]]) & ov[["volume_b"]] > 0
  has_ov <- !is.na(ov[["volume_overlap"]]) & ov[["volume_overlap"]] > 0
  categ <- terra::ifel(has_ov, 2,
    terra::ifel(has_a & has_b, 2,
      terra::ifel(has_a, 1,
        terra::ifel(has_b, 3, NA))))
  levels(categ) <- data.frame(
    id = 1:3,
    category = c(sp, "Intersect", fi)
  )

  ggplot() +
    geom_sf(data = bangladesh_eez, fill = "white", colour = "red",
            linewidth = 0.4, linetype = "dashed") +
    geom_sf(data = land_mask, fill = "grey90", colour = "grey50", linewidth = 0.5) +
    geom_spatraster(data = categ, show.legend = TRUE) +
    scale_fill_manual(
      values = setNames(
        c(col_species, col_intersect, col_fishery),
        c(sp, "Intersect", fi)
      ),
      limits = c(sp, "Intersect", fi),
      name = NULL,
      na.translate = FALSE,
      drop = FALSE
    ) +
    coord_sf(xlim = plot_bbox[c("xmin", "xmax")],
             ylim = plot_bbox[c("ymin", "ymax")], expand = FALSE) +
    theme_map() +
    theme(
      legend.position = "inside",
      legend.position.inside = c(0.7, 0.10),
      legend.background = element_rect(fill = "white", colour = NA),
      legend.key.size = unit(0.4, "cm"),
      legend.text = element_text(size = 8)
    )
})
names(overlap_map_plots) <- paste(cr_pairs$species, cr_pairs$fishery, sep = " x ")
```

```{r overlap-depth-plots}
# Depth profile line plot for each species x fishery pair
# Shows bathymetry cross-section with species/fishery depth ranges
overlap_depth_plots <- lapply(seq_len(nrow(cr_pairs)), function(i) {
  ov <- cr_overlap_rasters[[i]]

  # Stack overlap depth layers with bathymetry, extract all values at once
  stack <- c(ov[["depth_min_a"]], ov[["depth_max_a"]],
             ov[["depth_min_b"]], ov[["depth_max_b"]],
             ov[["depth_min_overlap"]], ov[["depth_max_overlap"]],
             seafloor)
  df <- as.data.frame(terra::values(stack)) %>%
    setNames(c("sp_depth_min", "sp_depth_max", "fi_depth_min", "fi_depth_max",
               "intersect_min", "intersect_max", "seafloor_depth")) %>%
    filter(!is.na(sp_depth_min) & !is.na(fi_depth_min)) %>%
    arrange(seafloor_depth) %>%
    mutate(cell_index = row_number())

  ggplot(df, aes(x = cell_index)) +
    geom_line(aes(y = -seafloor_depth), colour = "grey30") +
    geom_ribbon(aes(ymin = -sp_depth_max, ymax = -sp_depth_min),
                fill = col_species, alpha = 1) +
    geom_ribbon(aes(ymin = -fi_depth_max, ymax = -fi_depth_min),
                fill = col_fishery, alpha = 1) +
    geom_ribbon(aes(ymin = -intersect_max, ymax = -intersect_min),
                fill = col_intersect, alpha = 1) +
    annotate("text", x = max(df$cell_index) * 0.95, y = min(-df$seafloor_depth) * 0.95,
             label = paste0("n = ", max(df$cell_index)),
             hjust = 1, vjust = 0, size = 3) +
    labs(x = NULL, y = "Depth (m)") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.title = element_text(size = 8),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size = 7)
    )
})
names(overlap_depth_plots) <- paste(cr_pairs$species, cr_pairs$fishery, sep = " x ")
```

```{r overlap-combine}
# Combine depth sidebar + map for each pair (1/4 + 3/4 width)
overlap_plots <- lapply(seq_len(nrow(cr_pairs)), function(i) {
  overlap_depth_plots[[i]] + overlap_map_plots[[i]] +
    plot_layout(widths = c(1.5, 3)) &
    theme(plot.margin = margin(0, 0, 2, 0))
})
names(overlap_plots) <- paste(cr_pairs$species, cr_pairs$fishery, sep = " x ")

# Tile by species: 2 rows × 3 columns with collected legends
cr_species_4d <- unique(cr_pairs$species)
n_cols <- 2

out_dir_ov <- here("output", "bangladesh-fisheries-3d-overlap", "species_fishery_overlap")
dir.create(out_dir_ov, recursive = TRUE, showWarnings = FALSE)

for (sp in cr_species_4d) {
  sp_idx <- which(cr_pairs$species == sp)
  sp_plots <- overlap_plots[sp_idx]
  n_rows <- ceiling(length(sp_plots) / n_cols)

  tiled <- wrap_plots(sp_plots, ncol = n_cols, nrow = n_rows)

  fname <- gsub(" ", "_", sp)
  ggsave(file.path(out_dir_ov, paste0(fname, ".png")),
         tiled, width = 5 * n_cols, height = 6 * n_rows, dpi = 300)
}
```

## Conclusion

The raster-based approach in `sharkabc3d` generalises the hexagonal grid
method and leverages `terra`'s optimised raster algebra, making the analysis
faster and applicable to any region, species set, or fishery configuration.
