---
title: "Depth-Stratified Fishing Effort from Global Fishing Watch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Depth-Stratified Fishing Effort from Global Fishing Watch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Depth-Stratified Fishing Effort from Global Fishing Watch

This vignette walks through turning Global Fishing Watch (GFW) apparent
fishing-effort data from a flat, 2D per-cell product into a **3D,
depth-stratified effort stack** that can be intersected with shark and ray
ranges via [calc_volume_overlap()].

Ingest is delegated to the [gfwr](https://globalfishingwatch.github.io/gfwr/)
package, which queries the GFW 4Wings API and returns effort already
aggregated per cell × group. `sharkabc3d` covers the two steps gfwr does not:
rasterising onto the canonical study grid, and extending the 2D effort
product into a 3D depth-stratified stack via gear-class depth priors and
bathymetry. Both steps are implemented — `gfw_effort_to_raster()` and
`gfw_gear_depth_bands()` — but chunks here are `eval = FALSE` because they
need a `GFW_TOKEN` and a local bathymetry grid.

## What GFW provides, and what we need

GFW (Kroodsma et al. 2018, *Science*) publishes apparent fishing effort in
fishing hours per grid cell × time × flag state × gear type. It is a **2D
raster product**: a longline set hooks at 80 m and a bottom trawl scraping
the seafloor at 400 m both appear as undifferentiated "effort" pixels. For
species with strong vertical habitat partitioning, flat overlap over- or
under-counts risk.

The workflow has three steps:

1. **Fetch** effort for a region with `gfwr::gfw_ais_fishing_hours()`,
   grouped by gear type (and optionally flag).
2. **Rasterise** the returned tibble onto a target grid, one layer per
   group level, with [gfw_effort_to_raster()].
3. **Extend into 3D** by attaching a depth window per gear class and
   clamping benthic gears to bathymetry, producing a stack indexed by
   `effort_<geartype>_depth=<value>` with [gfw_gear_depth_bands()].

## Setup

```{r setup}
# library(sharkabc3d)
library(terra)
library(here)
library(gfwr)
```

`gfwr` requires an API token. Register at
<https://globalfishingwatch.org/our-apis/tokens> and put the key in the
`GFW_TOKEN` environment variable; `gfwr::gfw_auth()` will pick it up.

## Step 1: Fetch apparent fishing effort

`gfw_ais_fishing_hours()` returns a long-format tibble with one row per
cell × time bucket × group level. For an EEZ, MPA, or RFMO region, first
resolve the human-readable name to the numeric region ID with
`gfwr::gfw_region_id()` — the API expects the ID, not an ISO3 string.

```{r fetch-effort}
# Resolve the Japan EEZ to its numeric region ID.
jpn_eez <- gfw_region_id(region = "JPN", region_source = "EEZ")$id

effort <- gfw_ais_fishing_hours(
  spatial_resolution  = "HIGH",        # 0.01 deg
  temporal_resolution = "YEARLY",      # one bucket per year in [start, end]
  start_date          = "2022-01-01",
  end_date            = "2022-12-31",
  region              = jpn_eez,
  region_source       = "EEZ",
  group_by            = "GEARTYPE"
)

head(effort)
#> # A tibble: 6 x 6
#>     Lat   Lon `Time Range` geartype           `Vessel IDs` `Apparent Fishing Hours`
#>   <dbl> <dbl>        <dbl> <chr>                     <dbl>                    <dbl>
#> 1  35.4 140.1         2022 trawlers                      1                     0.95
#> 2  34.6 138.9         2022 inconclusive                  1                     1.85
#> 3  35.1 139.7         2022 fishing                       1                     3.05
#> 4  33.2 137.4         2022 drifting_longlines            1                     2.38
#> ...
```

Note: GFW returns gear-class labels in lower-case (`geartype` column,
values like `trawlers`, `drifting_longlines`). It also emits an
`inconclusive` class for vessels the classifier could not assign — these
are passed through and treated like `fishing` (unclassified) in the
depth-band logic.

## Step 2: Rasterise to a per-gear stack on the study grid

Pick a target grid — typically the same grid used for species ranges and
WOA extraction — and rasterise the effort tibble onto it. One SpatRaster
layer is produced per unique level of `layer_by`.

```{r effort-to-raster}
# Use the bathymetry grid as the canonical study grid.
bathy <- load_gebco_bathymetry(
  here("/home/jay/Programming_Projects/Big_Data/gebco_2025_sub_ice_topo/GEBCO_2025_sub_ice.nc")
)

# Crop bathymetry to the effort footprint (with a small buffer) before
# any heavy raster ops. GEBCO is global at 15 arc-seconds; cropping first
# turns a multi-GB reproject/aggregate into a few-MB one.
effort_extent <- terra::ext(
  min(effort$Lon) - 0.5, max(effort$Lon) + 0.5,
  min(effort$Lat) - 0.5, max(effort$Lat) + 0.5
)
bathy_crop <- terra::crop(bathy, effort_extent)

# Canonical study grid: positive-down depth on the effort footprint,
# coarsened to ~0.04 deg for the worked example.
study_grid <- (terra::project(bathy_crop, "EPSG:4326") * -1) |>
  terra::aggregate(fact = 10)

effort_by_gear <- gfw_effort_to_raster(
  effort   = effort,
  grid     = study_grid,
  layer_by = "geartype",
  value    = "Apparent Fishing Hours",
  fun      = "sum"
)

names(effort_by_gear)
#> [1] "effort_drifting_longlines" "effort_set_gillnets"
#> [3] "effort_trawlers"           "effort_pots_and_traps"
#> ...
```

If you need a different cut (e.g. one layer per flag, or per
flag × geartype), re-fetch with the appropriate `group_by` on the API —
that's cheaper and more honest than collapsing client-side.

## Step 3: Extend into 3D with a user-supplied gear-depth lookup

[gfw_gear_depth_bands()] needs a `depth_lookup` table mapping each GFW
gear class to an operating depth band. **The package does not ship a
default lookup**: operating depths vary by region, fleet, and time, and
the right values for a given analysis depend on assumptions the analyst
is responsible for. The example below is a placeholder used purely to
demonstrate the function's mechanics — it is *not* a recommendation for
real analyses. Replace it with literature- or fishery-specific values
appropriate to your study.

The lookup has five columns:

- `geartype` — the GFW gear class label (matching the `geartype` values
  in the effort tibble).
- `depth_min`, `depth_max` — shallowest and deepest operating depths in
  metres, positive-down. Set both to `NA` for benthic gears whose band
  is computed from bathymetry, and for unknown/unclassified gears.
- `mode` — `"pelagic"` (fixed band in the water column), `"benthic"`
  (band clamped to bathymetry ± `benthic_buffer`), or `"unknown"`
  (handled via the `fallback` argument).
- `benthic_buffer` — for `mode = "benthic"`, metres above the seafloor
  the gear is assumed to fish. `NA` otherwise.

```{r depth-lookup}
# Placeholder values for demonstration only — substitute literature- or
# fishery-specific operating depths for real analyses.
depth_lookup <- data.frame(
  geartype = c(
    "drifting_longlines", "set_longlines", "set_gillnets",
    "trawlers", "dredge_fishing", "pole_and_line",
    "trollers", "purse_seines", "tuna_purse_seines",
    "other_purse_seines", "seiners", "squid_jigger",
    "pots_and_traps", "fixed_gear", "fishing", "inconclusive"
  ),
  depth_min      = c(  0, NA,   0, NA, NA,   0,   0,   0,   0,   0,   0,   0, NA, NA, NA, NA),
  depth_max      = c(400, NA, 140, NA, NA,  20, 150, 200, 250, 100, 200, 100, NA, NA, NA, NA),
  mode = c(
    "pelagic", "benthic", "pelagic",
    "benthic", "benthic", "pelagic",
    "pelagic", "pelagic", "pelagic",
    "pelagic", "pelagic", "pelagic",
    "benthic", "benthic", "unknown", "unknown"
  ),
  benthic_buffer = c( NA, 50,  NA, 50, 10,  NA,  NA,  NA,  NA,  NA,  NA,  NA, 50, 50, NA, NA),
  stringsAsFactors = FALSE
)
```

Pelagic gears use a fixed depth band. Benthic gears ride the seafloor:
the band becomes `[max(bathy − buffer, 0), bathy]` per cell, handled by
[gfw_gear_depth_bands()]. The function processes one gear at a time so
peak memory stays bounded by a single gear's depth-stratified stack
rather than every gear at once. Loop over gears externally and concatenate
the results — writing each per-gear stack to a tempfile in the loop body
keeps the working set on disk for large analyses.

```{r gear-depth-bands}
# Use WOA's 57 standard depths (0–1500 m) as the vertical grid.
standard_depths <- c(
  0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90, 95, 100,
  125, 150, 175, 200, 225, 250, 275, 300, 325, 350, 375, 400, 425, 450, 475, 500,
  550, 600, 650, 700, 750, 800, 850, 900, 950, 1000,
  1050, 1100, 1150, 1200, 1250, 1300, 1350, 1400, 1450, 1500
)

# `study_grid` already carries positive-down depth on the cropped
# footprint; clamp non-ocean cells to 0 to get the seafloor reference.
seafloor <- terra::clamp(study_grid, lower = 0)

# Build a depth-stratified stack for one mode (benthic or pelagic). The
# per-gear loop writes each stack to a tempfile so peak memory stays
# bounded by one gear at a time — essential at Japan-EEZ scale.
build_mode_stack <- function(mode) {
  gears <- intersect(
    sub("^effort_", "", names(effort_by_gear)),
    depth_lookup$geartype[depth_lookup$mode == mode]
  )
  per_gear <- lapply(gears, function(gear) {
    s <- gfw_gear_depth_bands(
      effort_layer    = effort_by_gear[[paste0("effort_", gear)]],
      gear            = gear,
      bathymetry      = seafloor,
      standard_depths = standard_depths,
      depth_lookup    = depth_lookup,
      allocation      = "uniform"
    )
    terra::writeRaster(s, tempfile(fileext = ".tif"), overwrite = TRUE)
  })
  do.call(c, per_gear)
}

# Two stacks — benthic gears ride the seafloor, pelagic gears occupy a
# fixed depth band — kept separate so the visualisation can render them
# with distinct colour scales.
benthic_3d <- build_mode_stack("benthic")
pelagic_3d <- build_mode_stack("pelagic")

names(benthic_3d)[1:5]
#> [1] "effort_trawlers_depth=0"
#> [2] "effort_trawlers_depth=5"
#> [3] "effort_trawlers_depth=10"
#> [4] "effort_trawlers_depth=15"
#> [5] "effort_trawlers_depth=20"
```

## Visualising the 3D benthic-effort stack

A 3D scene is the most direct way to read `benthic_3d`: the cubes sit on
the seafloor, you can see whether effort is concentrated on the shelf or
in deeper canyons, and the bathymetric surface gives every cube a spatial
reference. We render with `plotly` so the scene stays interactive in the
rendered HTML vignette. Pelagic effort is excluded — those gears occupy
dozens of standard depths per cell and the cube count crashes plotly's
JSON serializer.

```{r 3d-visualization}
library(plotly)
library(tidyr)
library(dplyr)
library(stringr)

# `benthic_3d` and `pelagic_3d` are both produced by the gear-depth-bands
# chunk above; we render them as two mesh3d traces with distinct colour
# scales so the modes read as separate categories.

# Vertical extent of each depth band: half-way to neighbouring standard
# depths (so adjacent cubes meet without gaps or overlap).
sd_sorted <- sort(unique(standard_depths))
n_sd      <- length(sd_sorted)
mids      <- (sd_sorted[-n_sd] + sd_sorted[-1]) / 2
bands <- data.frame(
  depth  = sd_sorted,
  z_low  = c(max(sd_sorted[1] - (sd_sorted[2] - sd_sorted[1]) / 2, 0), mids),
  z_high = c(mids, sd_sorted[n_sd] +
                   (sd_sorted[n_sd] - sd_sorted[n_sd - 1]) / 2)
)

# Concatenate one mesh3d trace covering all cubes. Each cube has 8 vertices
# and 12 triangles; vertex indices in plotly's mesh3d are 0-indexed.
make_cube_mesh <- function(voxels, dx, dy) {
  n <- nrow(voxels)
  off <- as.matrix(expand.grid(sx = c(-1, 1), sy = c(-1, 1), sz = c(0, 1)))
  rows <- voxels[rep(seq_len(n), each = 8), ]
  o    <- off[rep(seq_len(8), times = n), ]

  vx <- rows$x + o[, "sx"] * dx
  vy <- rows$y + o[, "sy"] * dy
  vz <- ifelse(o[, "sz"] == 0, rows$z_low, rows$z_high)

  tris <- rbind(
    c(0, 1, 2), c(1, 3, 2),  # shallow face
    c(4, 5, 6), c(5, 7, 6),  # deep face
    c(0, 4, 1), c(1, 4, 5),  # south face
    c(2, 3, 6), c(3, 7, 6),  # north face
    c(0, 2, 4), c(2, 6, 4),  # west face
    c(1, 5, 3), c(3, 5, 7)   # east face
  )
  base_v <- rep(seq(0L, by = 8L, length.out = n), each = 12)
  list(
    x = vx, y = vy, z = vz,
    i = rep(tris[, 1], n) + base_v,
    j = rep(tris[, 2], n) + base_v,
    k = rep(tris[, 3], n) + base_v,
    intensity = rows$effort
  )
}

# Turn a multi-layer effort stack into mesh3d cubes by collapsing across
# gears at each (x, y, depth) and joining per-depth z-band extents.
stack_to_cubes <- function(stack, dx, dy, bands) {
  voxels <- terra::as.data.frame(stack, xy = TRUE, na.rm = FALSE) |>
    pivot_longer(cols = -c(x, y), names_to = "layer", values_to = "effort") |>
    filter(!is.na(effort) & effort > 0) |>
    mutate(depth = as.numeric(str_extract(layer, "(?<=depth=)-?[0-9.]+"))) |>
    group_by(x, y, depth) |>
    summarise(effort = sum(effort), .groups = "drop") |>
    left_join(bands, by = "depth")
  make_cube_mesh(voxels, dx = dx, dy = dy)
}

# Only render benthic effort. Pelagic gears occupy dozens of standard
# depths per cell, so the cube count blows past plotly's JSON serializer
# (R's 2 GB string limit) even after threshold-thinning. `pelagic_3d` is
# still produced by the gear-depth-bands chunk for the species-overlap
# analysis below.
res           <- terra::res(study_grid)
benthic_cubes <- stack_to_cubes(benthic_3d, res[1] / 2, res[2] / 2, bands)

# Log-scale colour mapping. Plotly's mesh3d colourscale is linear in
# `intensity`, so we feed it log10(effort) directly, then relabel the
# colourbar ticks back to the original-scale values (1, 10, 100, ...).
benthic_log_intensity <- log10(benthic_cubes$intensity)
log_breaks <- 10 ^ seq(floor(min(benthic_log_intensity)),
                       ceiling(max(benthic_log_intensity)))

# Bathymetry surface and a flat sea surface at z = 0. Plotly expects the
# i-th row of z to align with y[i], so flip both together south-to-north.
bathy_x   <- terra::xFromCol(seafloor, seq_len(terra::ncol(seafloor)))
bathy_y   <- terra::yFromRow(seafloor, seq_len(terra::nrow(seafloor)))
bathy_mat <- terra::as.matrix(seafloor, wide = TRUE)
bathy_mat <- bathy_mat[nrow(bathy_mat):1, ]
bathy_y   <- rev(bathy_y)
sea_z     <- matrix(0, nrow = length(bathy_y), ncol = length(bathy_x))


# Coastline: contour the *native-resolution* `bathy_crop` (positive
# elevation on land, negative below sea level) at 0 — the actual coast.
# Using `seafloor` instead loses the coast geometry: aggregation by
# fact=10 flattens 10×10 GEBCO cells, and clamp(lower = 0) snaps
# negative-elevation cells onto the same plane as sea-level cells, so
# the contour zigzags through the coarsened grid rather than tracing
# real shorelines. Project onto the same CRS as the plot if needed.
coast_lines <- terra::as.contour(bathy_crop, levels = 0)
if (!terra::same.crs(coast_lines, study_grid)) {
  coast_lines <- terra::project(coast_lines, terra::crs(study_grid))
}
# Flatten each line segment into one (x, y) sequence and separate
# segments with NAs so plotly's scatter3d breaks the line at NA,
# preserving disconnected coastline pieces (islands, archipelagos).
coast_xy    <- sf::st_coordinates(sf::st_as_sf(coast_lines))
coast_split <- split(seq_len(nrow(coast_xy)),
                     paste(coast_xy[, "L1"], coast_xy[, "L2"]))
coast_x <- unlist(lapply(coast_split, function(idx) c(coast_xy[idx, "X"], NA)))
coast_y <- unlist(lapply(coast_split, function(idx) c(coast_xy[idx, "Y"], NA)))
coast_z <- ifelse(is.na(coast_x), NA_real_, 0)

# Namespace-qualify every plotly call: `terra` (loaded earlier in this
# vignette) exports a `layout` generic, and `graphics::layout` is on the
# search path too. Either can mask `plotly::layout` and trigger
# "unused argument (scene = ...)".
build_benthic_scene <- function(camera = NULL) {
  scene <- list(
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude"),
    zaxis = list(title = "Depth (m)", autorange = "reversed"),
    # Compress the z-axis so depth doesn't dominate the scene. Tune the
    # `z` value down for less vertical stretch, up for more.
    aspectmode  = "manual",
    aspectratio = list(x = 1, y = 1, z = 0.05)
  )
  if (!is.null(camera)) scene$camera <- camera

  plotly::plot_ly() |>
    plotly::add_surface(
      x = bathy_x, y = bathy_y, z = bathy_mat,
      # Single surface keeps the coast geometry continuous — splitting
      # land/ocean into two NA-masked surfaces drops faces that bridge
      # the boundary, leaving ragged edges. Plotly interpolates colours
      # across faces, so the green→tan transition smears across the
      # 1-cell-wide coastal band; the coastline scatter trace below
      # overlays a sharp dark line that visually anchors the boundary.
      colorscale = list(
        c(0,    "rgb(140, 170, 110)"),  # land (z = 0): muted green
        c(1e-4, "rgb(180, 170, 150)"),  # shallow ocean: tan
        c(1,    "rgb( 60,  45,  30)")   # deep ocean: dark brown
      ),
      cmin = 0,
      cmax = max(bathy_mat, na.rm = TRUE),
      showscale = FALSE, opacity = 1, name = "Seafloor"
    ) |>
    plotly::add_trace(
      type = "scatter3d", mode = "lines",
      x = coast_x, y = coast_y, z = coast_z,
      line = list(color = "rgb(30, 25, 20)", width = 3),
      hoverinfo = "skip", showlegend = FALSE, name = "Coastline"
    ) |>
    plotly::add_trace(
      type = "mesh3d",
      x = benthic_cubes$x, y = benthic_cubes$y, z = benthic_cubes$z,
      i = benthic_cubes$i, j = benthic_cubes$j, k = benthic_cubes$k,
      intensity = benthic_log_intensity,
      colorscale = "Viridis", opacity = 0.5, flatshading = TRUE,
      colorbar = list(
        title    = "Effort\n(hours)",
        tickvals = log10(log_breaks),
        ticktext = format(log_breaks, big.mark = ",", scientific = FALSE,
                          trim = TRUE)
      ),
      name = "Benthic effort"
    ) |>
    plotly::layout(scene = scene)
}

build_benthic_scene()
```

**Figure 1.** Depth-stratified benthic fishing effort in the Japanese EEZ
(2022), rendered as an interactive 3D scene. Each cube is one study-grid cell
spanning one World Ocean Atlas standard depth band, coloured by apparent
fishing effort in hours — summed across benthic gear types — on a log₁₀ scale.
The shaded surface is GEBCO bathymetry, from muted green land through tan
shelf to dark brown deep water, and the dark line traces the coastline. The
depth axis is reversed so the sea surface sits at the top and the seafloor at
the bottom.

## Exporting an animation for slides

Plotly widgets are interactive in HTML but PowerPoint can't embed them.
The path below renders the same scene at a series of camera angles
(turntable rotation), captures each frame as a PNG via `webshot2`, and
stitches the frames into an MP4 with `av` (which wraps ffmpeg). MP4
embeds natively in PowerPoint and Keynote.

`webshot2` needs a Chrome/Chromium install (one-time:
`webshot2::install_chrome()`). `av` pulls ffmpeg via the bundled binary
on most platforms, no system install required.

```{r 3d-animation}
library(webshot2)
library(av)

# Turntable parameters. Plotly's camera works in scene-coordinate units;
# `eye` is the camera position relative to the scene's centre. Distance
# from origin is sqrt(x^2 + y^2 + z^2) ≈ 1.8 here. Adjust `elev_z` for
# pitch (higher = more top-down) and `radius` for zoom.
n_frames <- 60
radius   <- 1.7
elev_z   <- 0.6
angles   <- seq(0, 2 * pi, length.out = n_frames + 1)[-(n_frames + 1)]

frame_dir   <- tempfile("benthic_frames_")
dir.create(frame_dir)
frame_files <- character(n_frames)

for (i in seq_along(angles)) {
  theta <- angles[i]
  cam <- list(
    eye    = list(x = radius * cos(theta), y = radius * sin(theta), z = elev_z),
    center = list(x = 0, y = 0, z = 0),
    up     = list(x = 0, y = 0, z = 1)
  )
  fig <- build_benthic_scene(camera = cam)

  html_path <- tempfile(fileext = ".html")
  png_path  <- file.path(frame_dir, sprintf("frame_%03d.png", i))
  htmlwidgets::saveWidget(fig, html_path, selfcontained = TRUE)
  # `delay` gives the WebGL canvas time to render before screenshot.
  webshot2::webshot(html_path, png_path,
                    vwidth = 1280, vheight = 720, delay = 1.5)
  frame_files[i] <- png_path
}

# 24 fps × 60 frames = 2.5 s loop. Bump n_frames or fps for smoother
# rotation; keep the file under ~10 MB for easy slide embedding.
out_mp4 <- here("figures/benthic_effort_3d.mp4")
av::av_encode_video(frame_files, output = out_mp4, framerate = 24)

# Optional GIF fallback for venues that block MP4 (rarer).
# magick::image_animate(magick::image_read(frame_files), fps = 24) |>
#   magick::image_write(here("presentation/benthic_effort_3d.gif"))
```

Drop the resulting `.mp4` straight into a slide via
**Insert → Video → Video on My PC** (PowerPoint) or **Insert → Movie**
(Keynote). PowerPoint will play it in-place; for a presentation-mode
auto-loop, set the playback options to *Start: Automatically* and
*Loop until Stopped*.

## Intersecting with a species' 3D range

Once `benthic_3d` / `pelagic_3d` follow the `{variable}_depth=<value>`
convention, they compose with the rest of the package the same way WOA
rasters do:

```{r species-overlap}
# Example: silky shark (epipelagic, shallow exposure to longlines).
# The IUCN shark/ray/chimaera shapefile carries one row per species range
# polygon. Filter at read time with SQL so the in-memory object is small.
iucn_shp <- "/home/jay/Programming_Projects/Big_Data/SHARKS_RAYS_CHIMAERAS/SHARKS_RAYS_CHIMAERAS.shp"

sp_range <- sf::st_read(
  iucn_shp,
  query = "SELECT * FROM SHARKS_RAYS_CHIMAERAS WHERE sci_name = 'Carcharhinus falciformis'",
  quiet = TRUE
)

# Depth limits are placeholder values for demonstration; in practice pull
# them per-species via fetch_species_assessments() / fill_missing_depths()
# from the IUCN Red List API.
range_rast <- vect_to_envelope(
  polygon   = sp_range,
  template  = study_grid,
  depth_min = 0,
  # The seafloor is just another depth_max constraint: per cell the shallowest
  # of the two wins, so the range never extends below the bed.
  depth_max = list(500, seafloor)
)

# Mask the 3D effort stack to cells + depths inside the species' per-cell
# depth window. extract_rast_range() reads the `effort_<gear>_depth=<value>`
# layer names and keeps each layer's effort only where the species is
# present at that depth.
longline_layers <- pelagic_3d[[grep("drifting_longlines", names(pelagic_3d))]]
exposed_effort  <- extract_rast_range(range_rast, longline_layers)

# Total longline effort-hours overlapping the silky shark's 3D range.
total_exposed <- sum(
  terra::global(exposed_effort, "sum", na.rm = TRUE)$sum,
  na.rm = TRUE
)
total_exposed
#> [1] <hours>

# Per-depth exposure profile (effort-hours summed across cells, per layer).
per_depth <- terra::global(exposed_effort, "sum", na.rm = TRUE)
per_depth$depth <- as.numeric(
  stringr::str_extract(rownames(per_depth), "(?<=_depth=)-?[0-9.]+")
)
per_depth[!is.na(per_depth$sum) & per_depth$sum > 0,
          c("depth", "sum")]
```

## Caveats worth surfacing

These split into two groups: limitations of the underlying GFW AIS effort
product (sourced from
<https://globalfishingwatch.org/data-documentation/apparent-fishing-effort-ais/>),
and limitations of the depth-stratification step this package adds on top.

### Inherited from GFW AIS effort

- **AIS coverage skews to large, offshore vessels.** AIS carriage is
  primarily mandated for vessels over 24 m operating offshore; small-scale
  and artisanal fleets are largely absent from the dataset. SAR
  dark-vessel detections (Paolo et al. 2024) are the complementary layer
  for artisanal-fleet coverage.
- **Class A vs. Class B reception is uneven.** Smaller AIS-equipped
  vessels carry Class B units, which transmit at lower power and are
  detected less reliably by satellites — so smaller-vessel effort is
  systematically under-represented even within the AIS-equipped fleet.
- **AIS can be disabled or spoofed.** Deliberate gaps and falsified
  identities produce missing or misattributed positions.
- **Signal interference in busy areas.** In high-traffic regions
  overlapping AIS transmissions degrade satellite detection, reducing
  apparent effort.
- **"Apparent" fishing is a behavioural inference.** The classifier
  flags movement patterns consistent with searching, setting,
  soaking, and hauling — not just confirmed gear deployment. Expect
  false positives where vessels slow or change course for non-fishing
  reasons.
- **Vessel-type and gear misclassification.** Registry data is
  inconsistent; MMSI recycling can mix two vessels' tracks; vessels that
  switch gears are inherently hard to classify. Treat per-gear effort as
  a best-available label, not a measurement.
- **Temporal comparisons are confounded by AIS adoption.** Year-on-year
  changes in apparent effort partly reflect changes in AIS uptake,
  satellite coverage, and terrestrial-station access, not just changes
  in real fishing activity.

### Added by this package's depth-stratification

- **The depth lookup is the analyst's responsibility.** The package does
  not ship a default `depth_lookup`. GFW does not publish set depth, so
  any per-gear band is an assumption — and the right assumption depends
  on the region, fleet, and time period being analysed.
- **Pelagic vs. benthic variants of gillnets and trawls are not always
  distinguished** in GFW's classification. The midwater/bottom trawl
  case is the clearest example: in 2D the distinction is invisible, but
  in 3D the two gears live in completely different parts of the water
  column. Whatever you assign in `depth_lookup` will carry that
  uncertainty until GFW separates them.
- **`allocation = "uniform"` preserves total effort-hours** but assumes
  gear distributes evenly across its band; `"presence"` is cheaper but
  double counts and is only appropriate for footprint maps.

## Function summary

| Function | Purpose |
|---|---|
| `gfwr::gfw_ais_fishing_hours()` | Fetch GFW apparent fishing effort for a region, aggregated server-side by `group_by` |
| [gfw_effort_to_raster()] | Rasterise the gfwr tibble onto the study grid, one layer per group level |
| [gfw_gear_depth_bands()] | Extend a per-gear effort raster into a 3D depth-stratified stack using a user-supplied gear → depth-band lookup |
