---
title: "Extracting 3D Environmental Data for a Single Species"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting 3D Environmental Data for a Single Species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Extracting 3D Environmental Data for a Single Species

This vignette is a scoped-down version of the full WOA environmental
extraction workflow that runs for **one example species** instead of every
shark and ray. It produces two visualisation products:

1. An **animated GIF** of annual mean temperature through the species' range
   across every depth layer it occupies.
2. A **three-panel static figure** (stacked vertically) showing temperature at
   the surface, at the standard depth closest to the halfway point, and at the
   deepest depth the species is found at.

The example species is the Banded Wobbegong (*Orectolobus halei*).

## Setup

```{r setup}
library(sharkabc3d)
library(dplyr)
library(sf)
library(stringr)
library(terra)
library(here)
library(ggplot2)
library(gganimate)
library(tidyr)
library(patchwork)
library(tidyterra)

# The single example species.
sp_name <- "Orectolobus halei"
```

## Step 1: Fetch depth limits for the example species

Fetching by `species_names` returns just this one species' assessment, so
there is no 30-minute group download.

```{r fetch-depths}
api_key <- Sys.getenv("IUCN_REDLIST_KEY")

depth_table <- fetch_species_assessments(
  api_key = api_key,
  species_names = sp_name
)

# Fill missing depths (a no-op if the species already has both limits).
depth_table_complete <- depth_table |>
  mutate(fill_missing_depths(upper_depth_limit, lower_depth_limit, genus_name)) |>
  filter(!is.na(upper_depth), !is.na(lower_depth))

sp_depths <- depth_table_complete[
  depth_table_complete$scientific_name == sp_name, ]
sp_id <- sp_depths$sis_id
```

## Step 2: Load the species range polygon

Only the one species is read out of the IUCN shapefile via SQL filtering.

```{r load-range}
iucn_shp <- here("/home/jay/Programming_Projects/Big_Data/SHARKS_RAYS_CHIMAERAS/SHARKS_RAYS_CHIMAERAS.shp")
layer_name <- sf::st_layers(iucn_shp)$name[1]

sp_range <- sf::st_read(
  iucn_shp,
  query = str_glue(
    "SELECT * FROM \"{layer_name}\" ",
    "WHERE id_no = {sp_id} ",
    "AND presence IN (1) AND origin IN (1, 2) AND seasonal IN (1, 2, 3)"
  ),
  quiet = TRUE
)
```

## Step 3: Prepare the WOA temperature raster and study grid

```{r prepare-rasters}
# Annual mean temperature (0.25 deg) is both the environmental raster and the
# canonical study grid.
t_annual_path <- woa_download("temperature", period = "annual", resolution = "0.25")
t_annual <- woa_load_nc(t_annual_path, field = "an")

# Crop to the species range (plus a ~1 degree margin) before any heavy work.
# WOA ships globally, but the bathymetry projection, voxelization, and
# per-depth masking all cost in proportion to grid size. A circumglobal range
# whose bbox spans the antimeridian makes this a no-op.
range_ext <- terra::intersect(
  terra::ext(terra::vect(sp_range)) + 1,
  terra::ext(-180, 180, -90, 90)
)
t_annual <- terra::crop(t_annual, range_ext)

# Derive the study grid from the cropped raster so the two stay aligned, as
# extract_rast_range() requires.
study_grid <- t_annual[[1]]

# Bathymetry -> positive depth, aligned to the study grid.
# Project before negating: negating first forces an eager pass over the full
# global 15 arc-second grid, which terra otherwise avoids by reading only the
# window the target grid needs.
bathy <- load_gebco_bathymetry("/home/jay/Programming_Projects/Big_Data/gebco_2025_sub_ice_topo/GEBCO_2025_sub_ice.nc")
seafloor <- (terra::project(bathy, study_grid) * -1) |>
  terra::clamp(lower = 0)
```

## Step 4: Voxelize the species range

`voxelize_range()` builds the per-cell depth window, clamping the species'
nominal depth limits to the seafloor at each cell.

```{r voxelize}
range_rast <- voxelize_range(
  polygons = sp_range,
  voxel = study_grid,
  bathymetry = seafloor,
  depth_min = sp_depths$upper_depth,
  depth_max = sp_depths$lower_depth
)

# Mask temperature to the cells + depths the species actually occupies. The
# grid is already scoped to the range, and cells outside the polygon are NA,
# so no further cropping is needed.
masked <- extract_rast_range(range_rast, t_annual)
```

## Figure 1: Animated GIF across all depths

Each frame is one depth layer, from the surface down to the deepest layer the
species occupies. The colour scale is locked to the global temperature range so
colours are comparable across frames.

```{r figure-1-animation}
long_df <- terra::as.data.frame(masked, xy = TRUE, na.rm = FALSE) |>
  pivot_longer(cols = -c(x, y), names_to = "layer", values_to = "value") |>
  mutate(depth = as.numeric(str_extract(layer, "(?<=_depth=)-?[0-9.]+"))) |>
  filter(!is.na(value))

# Lock the legend to the global min/max so colour mapping is identical per frame.
v_limits <- range(long_df$value, na.rm = TRUE)

# Factor with numerically-sorted levels so frames play 0 m -> deepest.
long_df$depth <- factor(long_df$depth, levels = sort(unique(long_df$depth)))

p_anim <- ggplot(long_df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  geom_sf(data = sp_range, fill = NA, colour = "black",
          linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "Temp (°C)", limits = v_limits,
                       option = "turbo",
                       na.value = "transparent") +
  coord_sf() +
  labs(subtitle = paste0(sp_name, " — depth: {current_frame} m")) +
  theme_minimal() +
  theme(
    plot.subtitle = element_text(face = "italic"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid.major = element_blank()
  ) +
  transition_manual(depth)

anim <- animate(
  p_anim,
  nframes = nlevels(long_df$depth),
  fps = 4, width = 700, height = 400,
  renderer = gifski_renderer()
)

anim_save(here("figures/single_species_depth_animation.gif"), anim)
anim
```

**Figure 1.** Annual mean sea temperature (World Ocean Atlas) within the
geographic range of the Banded Wobbegong (*Orectolobus halei*), animated
through every standard depth layer the species occupies. Each frame maps one
depth, from the sea surface (0 m) down to the deepest layer within the
species' bathymetry-clamped depth limits; the black outline is the IUCN
range polygon. The colour scale is fixed across all frames, so warming or
cooling with depth is directly comparable frame to frame.

## Figure 2: Three-panel static figure (surface / halfway / deepest)

We pick three depths from the layers the species actually occupies: the
surface (0 m), the standard WOA depth closest to the halfway point, and the
deepest layer with data inside the range.

```{r figure-2-depths}
# Depths of each masked layer, and which layers hold any data for this species.
layer_depths <- as.numeric(str_extract(names(masked), "(?<=_depth=)-?[0-9.]+"))
has_data <- vapply(seq_len(terra::nlyr(masked)), function(i) {
  !all(is.na(terra::values(masked[[i]])))
}, logical(1))
occupied_depths <- sort(layer_depths[has_data])

surface_depth <- 0
deepest_depth <- max(occupied_depths)
# Standard depth closest to the midpoint between surface and deepest occupied.
halfway_target <- (surface_depth + deepest_depth) / 2
halfway_depth <- occupied_depths[which.min(abs(occupied_depths - halfway_target))]

panel_depths <- c(surface_depth, halfway_depth, deepest_depth)
```

The `plot_range_at_depth()` helper below is defined in this article rather
than exported by the package — it is a thin `ggplot2` wrapper that draws one
map per depth. `patchwork` stacks them vertically.

```{r plot-helper}
# Map view of a species range with environmental values at the depth layer
# nearest `depth`.
plot_range_at_depth <- function(species_range, depth, rast_3d) {
  depths <- as.numeric(str_extract(names(rast_3d), "(?<=_depth=)-?[0-9.]+"))
  idx <- which.min(abs(depths - depth))
  layer <- rast_3d[[idx]]
  actual_depth <- depths[idx]

  range_vect <- if (inherits(species_range, "sf")) {
    terra::vect(species_range)
  } else {
    species_range
  }
  if (terra::crs(range_vect) != terra::crs(layer) &&
      terra::crs(range_vect) != "") {
    range_vect <- terra::project(range_vect, terra::crs(layer))
  }

  cropped <- terra::crop(layer, range_vect, mask = TRUE, touches = TRUE)
  range_sf <- if (inherits(species_range, "sf")) {
    species_range
  } else {
    sf::st_as_sf(range_vect)
  }

  ggplot() +
    geom_spatraster(data = cropped) +
    geom_sf(data = range_sf, fill = NA, colour = "black", linewidth = 0.3) +
    scale_fill_viridis_c(name = NULL, na.value = "transparent") +
    coord_sf() +
    labs(subtitle = paste0("Depth: ", actual_depth, " m")) +
    theme_minimal() +
    theme(axis.title = element_blank())
}
```

```{r figure-2-plot, fig.width=6, fig.height=12}
# Shared colour limits across the three panels. Without explicit `limits` each
# panel rescales to its own range, so the warm surface and the cold deepest
# layer would render in the same colours and the panels would not be
# comparable. Mirror plot_range_at_depth()'s own layer pick and crop so the
# limits cover exactly the values it draws.
panel_layers <- terra::rast(lapply(panel_depths, function(d) {
  t_annual[[which.min(abs(layer_depths - d))]]
}))
fill_limits <- range(
  terra::values(terra::crop(panel_layers, terra::vect(sp_range),
                            mask = TRUE, touches = TRUE)),
  na.rm = TRUE
)

panels <- lapply(seq_along(panel_depths), function(i) {
  d <- panel_depths[i]
  plot_range_at_depth(species_range = sp_range, depth = d, rast_3d = t_annual) +
    scale_fill_viridis_c(name = "Temp (°C)", na.value = "transparent",
                         option = "turbo", limits = fill_limits) +
    # Fold the panel letter into the subtitle rather than using patchwork's
    # tag_levels, which renders the tag on its own line above this text.
    labs(subtitle = paste0(LETTERS[i], ". Depth: ", d, " m"))
})

fig2 <- (panels[[1]] / panels[[2]] / panels[[3]]) +
  plot_annotation(
    title = sp_name,
    theme = theme(plot.title = element_text(face = "italic"))
  ) +
  plot_layout(guides = "collect")

ggsave(here("figures/single_species_three_panel.png"), fig2,
       width = 6, height = 12, dpi = 300)
fig2
```

**Figure 2.** Annual mean sea temperature (World Ocean Atlas) within the
range of the Banded Wobbegong (*Orectolobus halei*) at three depths, stacked
from shallow (top) to deep (bottom): **(A)** the sea surface (0 m), **(B)**
the standard WOA depth nearest the midpoint of the species' occupied depth
range, and **(C)** the deepest standard depth at which the species occurs
(its depth limit clamped to the seafloor). Each panel is masked to the IUCN
range polygon and shares a common temperature colour scale, illustrating how
the thermal environment available to the species cools with depth.
